---
title: "Landscape analysis example"
author: "Trond Simensen"
date: "03 11 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Landscape analysis example"
author: "Trond Simensen"
date: "31 10 2019"
output: html_document
---

### Analysis of variation of landscape properties 
# Coastal plains

All four axes in the 4-dimensional GNMDS ordination of the KS457 dataset were confirmed by DCA. GNMDS ordination had Procrustes SS = 0.0363 and seven unstable OUs (which switched positions between the two best GNMDS solutions).

```{r, include = FALSE}
rm(list=ls())
#setwd("C:/Users/trosim/Documents/PhD/Subproject_2/Landscape_Analysis_Git_Hub/NiN_Landscape_Analysis")
#getwd()

# Loading libraries
library (readxl)#for reading in data
library(vegan)  #vegan 2.4.0
library(graphics)  #for å kunne identifisere punkter på ordinasjonsdiagram
library(knitr)
#library(formattable)
library(knitr)
library(kableExtra)


#Importerer data(matriser)
#Aktuelle datamatriser er: 

spreadsheet_filename  <-  "KS457_Coastal_plains.xlsx" # Setting the filename to the spreadsheet containg data

# Reading in data, specifying which columns to read with the command 'range'
sheetname  <-  "KS457" # Setting the name of the sheet within the spreadsheet containing data

av <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("J:CO"), col_names = TRUE) #analysis variables, e.g. the 85 variables in the data; nod id's
av  <-  as.data.frame(av)

lg <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("C:H"), col_names = TRUE) # placement along CLGs (affiliation to tentative major type based on the Norland project)
lg  <-  as.data.frame(lg)

nv <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("CQ:CX"), col_names = TRUE) # key variables
nv <- as.data.frame(nv)

id <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("A:B"), col_names = TRUE) #ID's and area
id <- as.data.frame(id)

sheetname  <-  "OrdRes" # Setting the name of the sheet within the spreadsheet containing data
onv <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("O:S"), col_names = TRUE) #rONV'er
onv <- as.data.frame(onv)

sheetname  <-  "Resk2" # Setting the name of the sheet within the spreadsheet containing data
snv <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("B:K"), col_names = TRUE) #reskalerte rONV'er (jf. regneark /resk)
snv <- as.data.frame(snv)

sheetname  <-  "JAtype" # Setting the name of the sheet within the spreadsheet containing data
jk <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("F:G"), col_names = TRUE) #jordartsklasser

sheetname  <-  "OrdRes" # Setting the name of the sheet within the spreadsheet containing data
ordx <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("B:E"), col_names = TRUE) #DCA
ordx <- as.data.frame(ordx)

dca1 <- ordx[,1]
dca2 <- ordx[,2]
dca3 <- ordx[,3]
dca4 <- ordx[,4]

sheetname  <-  "OrdRes" # Setting the name of the sheet within the spreadsheet containing data
ordx <- read_excel(spreadsheet_filename, sheet = sheetname, range = cell_cols("K:N"), col_names = TRUE) #GNMDS
ordx <- as.data.frame(ordx)

gnmds1 <- ordx[,1]
gnmds2 <- ordx[,2]
gnmds3 <- ordx[,3]
gnmds4 <- ordx[,4]

gnmds4_1 <- gnmds1
gnmds4_2 <- gnmds2
gnmds4_3 <- gnmds3 
gnmds4_4 <- gnmds4

akse1<-gnmds4_1
akse2<-gnmds4_2
akse3<-gnmds4_3
akse4<-gnmds4_4

attach(av)
names(av)
attach(lg)
names(lg)
attach(nv)
names(nv)
attach(id)
names(id)
attach(onv)
names(onv)
attach(snv)
names(snv)
attach(jk)
names(jk)

JK1 <- factor(JK1)
JK2 <- factor(JK2)
```


```{r, include= FALSE}
##### DCA #####
par(mfrow=c(1,1))

#Enkel plotting av DCA-resultater, svarte prikker:
plot(dca1,dca2,xlab="DCA1 - Coastal plains (KS457)",ylab="DCA2 - Coastal plains (KS457)",type="n")
```


```{r, echo = FALSE}
plot(dca1,dca2,xlab="DCA1 - Coastal plains (KS457)",ylab="DCA2 - Coastal plains (KS457)",pch=16)

```

There were no visible artifacts in the ordination charts. The correlation analysis, the vector charts and the iso-line charts show a strong first axis related to land use (infrastructure and agricultural character) that appears to be conditioned by an underlying geo-ecological gradient from inner to outer coast. Second axis is strongly related to relief.

```{r, include = FALSE}
########################################################################
#### Tolkning av ordinasjon vha korrelasjonsanalyse og enkle tester ####
########################################################################

#### Velg akser som skal analyseres

GNMDS4_1  <-  gnmds1
GNMDS4_2  <-  gnmds2
GNMDS4_3  <-  gnmds3
GNMDS4_4  <-  gnmds4

#######Beregning av Kendall's tau mellom en ordinasjonsakse eller variabel og analysevariablene#######

ax <- gnmds1 #Bestem akse
env.var <- av #Bestem miljøvariabelmatrise

p.val <- NULL #create empty objects for the p.values
tau.val <- NULL #creates empty object for the t-values.
p <- NULL

for(i in 1:84) # A Loop that calculates Kendall's tau and p-values for all the correlations tests.
{
	z <- cor.test(ax,env.var[,i], method="k")
	p.val[i] <- round(z$p.value, digits=5)
	tau.val[i] <- round(z$estimate,digits=4)
	if (p.val[i] < 0.05) #prints a star when the p-value is less than 0.05
	{p[i] <- "###"
      	}else{
  	p[i] <- " "}
}
#p.val
#tau.val
(test.summary.ax = data.frame(cbind(names(env.var),p.val,tau.val,p)))
corr.matrise1 <- as.data.frame(test.summary.ax [,c(1,3)])
names(corr.matrise1) <- c("Variable", "GNMDS1" )

##
ax <- gnmds2 #Bestem akse
env.var <- av #Bestem miljøvariabelmatrise

p.val <- NULL #create empty objects for the p.values
tau.val <- NULL #creates empty object for the t-values.
p <- NULL

for(i in 1:84) # A Loop that calculates Kendall's tau and p-values for all the correlations tests.
{
	z <- cor.test(ax,env.var[,i], method="k")
	p.val[i] <- round(z$p.value, digits=5)
	tau.val[i] <- round(z$estimate,digits=4)
	if (p.val[i] < 0.05) #prints a star when the p-value is less than 0.05
	{p[i] <- "###"
      	}else{
  	p[i] <- " "}
}
#p.val
#tau.val
(test.summary.ax = data.frame(cbind(names(env.var),p.val,tau.val,p)))
corr.matrise2 <- as.data.frame(test.summary.ax [,3])
names(corr.matrise2) <- c("GNMDS2" )

##
ax <- gnmds3 #Bestem akse
env.var <- av #Bestem miljøvariabelmatrise

p.val <- NULL #create empty objects for the p.values
tau.val <- NULL #creates empty object for the t-values.
p <- NULL

for(i in 1:84) # A Loop that calculates Kendall's tau and p-values for all the correlations tests.
{
	z <- cor.test(ax,env.var[,i], method="k")
	p.val[i] <- round(z$p.value, digits=5)
	tau.val[i] <- round(z$estimate,digits=4)
	if (p.val[i] < 0.05) #prints a star when the p-value is less than 0.05
	{p[i] <- "###"
      	}else{
  	p[i] <- " "}
}
#p.val
#tau.val
(test.summary.ax = data.frame(cbind(names(env.var),p.val,tau.val,p)))
corr.matrise3 <- as.data.frame(test.summary.ax [,3])
names(corr.matrise3) <- c("GNMDS3" )

##
ax <- gnmds4 #Bestem akse
env.var <- av #Bestem miljøvariabelmatrise

p.val <- NULL #create empty objects for the p.values
tau.val <- NULL #creates empty object for the t-values.
p <- NULL

for(i in 1:84) # A Loop that calculates Kendall's tau and p-values for all the correlations tests.
{
	z <- cor.test(ax,env.var[,i], method="k")
	p.val[i] <- round(z$p.value, digits=5)
	tau.val[i] <- round(z$estimate,digits=4)
	if (p.val[i] < 0.05) #prints a star when the p-value is less than 0.05
	{p[i] <- "###"
      	}else{
  	p[i] <- " "}
}
#p.val
#tau.val
(test.summary.ax = data.frame(cbind(names(env.var),p.val,tau.val,p)))
corr.matrise4 <- as.data.frame(test.summary.ax [,3])
names(corr.matrise4) <- c("GNMDS4" )
###

#Combining values to a dataframe
corr.matrise <- cbind(corr.matrise1, corr.matrise2, corr.matrise3, corr.matrise4)
corr.matrise$GNMDS1 <-  as.numeric(levels(corr.matrise$GNMDS1)[corr.matrise$GNMDS1])
corr.matrise$GNMDS2 <-  as.numeric(levels(corr.matrise$GNMDS2)[corr.matrise$GNMDS2])
corr.matrise$GNMDS3 <-  as.numeric(levels(corr.matrise$GNMDS3)[corr.matrise$GNMDS3])
corr.matrise$GNMDS4 <-  as.numeric(levels(corr.matrise$GNMDS4)[corr.matrise$GNMDS4])
```

```{r, include= FALSE}
Description <- c("No commercial buildings", "No fisheries-related buildings", "PA built-up", "PA flat terrain", "PA regulated magazine", "PA road", "PA thick layer of till", "PA depressions", "PA convex terrain", "Terrain form TPI1, numerical, mean", "PA strong ocean current", "PA normal ocean current", "PA weak ocean current", "PA large river", "PA trail, path", "PA steep terrain/slope", "PA landslide soil", "No summer mountain pasture", "PA old buildings", "Terrain ruggedness VRM3, mean", "Altitudinal range", "PA lime-rich bedrock", "PA lime-poor bedrock geology", "PA reindeer husbandry facilities", "PA plutonic rock", "PA volcanic rock", "PA sedimentary rock", "PA metamorphic rock", "PA terrestrial area", "PA built-up area", "PA town/city area", "PA mire", "PA moderate slope", "No marine islands", "PA marine deposits", "PA power lines", "PA lacustrine deposits", "PA glaciofluvial deposits", "PA exposed bedrock", "PA freshwater lake", "No freshwater lake islands", "Hydrographic index, ER, mean", "PA river", "PA exposed coast", "PA sligyhtly exposed coast", "PA slightly protected coast", "No lakes", "No cultural heritage sites outdoors", "No technical heritage sites", "No marine cultural heritage sites", "No church ruins", "No ancient rock art sites", "No archeological heritage sites", "PA steep coast", "PA flat coast", "Coastal ruggedness, VRM3, mean", "Coastal ruggedness, coarse scale, VRM9, mean", "PA rugged coast", "PA smooth/flat coast", "Coastal complexity", "PA complex coastline", "PA simple coastline Gab_fi", "PA boreal heaths", "PA south facing terrain", "PA north facing terrain", "PA large buildings", "PA open areas", "PA coniferous forest", "PA mixed boreal forest", "PA arable land", "PA deciduous forest", "PA surface cultivated land", "Distance to mire, mean", "Distance to lake, mean", "Distance to coast, mean", "PA rugged terrain", "PA rugged terrain, TPI6", "PA impediment", "PA patchy open treeless area", "PA lichen heath", "PA dry heath/open areas", "PA moist/fresh heath/open areas", "PA unregistered heath/open areas", "Inverse island size")

corr.matrise <- cbind(corr.matrise, Description)
corr.matrise <- corr.matrise [ , c(1,6,2:5)]
```

```{r, echo=FALSE}
kable(corr.matrise) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", font_size = 9))
```


```{r, cache = TRUE, include=FALSE}
##################################################
#### Standardskript for ordinasjonsdiagrammer ####
##################################################

#### Generelt ordinasjonsdiagrammer med like enheter på begge aksene ####

#Bestemmer gradientlengdene for akse 1, 2 og 3 og 4
grl1 <- max(akse1)-min(akse1)
grl2 <- max(akse2)-min(akse2)
grl3 <- max(akse3)-min(akse3)
grl4 <- max(akse4)-min(akse4)
grl1
grl2
grl3
grl4

#Diagram for akse 1 mot akse 2
#Vil finne koordinatene for hjørner i det 'plotteområdet' med koordinater (x1, x2, y1, y2) som gir
#like enheter på begge akser. Vi holder x1, x2 og y1 faste og bestemmer y2 slik at (x2-x1)/(y2-y1) i plt=c(x1,x2,y1,y2)
#blir lik grl1/grl2
#Det krever at y2 = [(x2-x1)*grl2 + y1*grl1]/grl1 
#Setter vi x1, x2 og y1 konstante, får vi:
y2 <- (0.8*grl2+0.15*grl1)/grl1
y2

#Diagram for akse 1 mot akse 3
#Vil finne koordinatene for hjørner i det 'plotteområdet' med koordinater (x1, x2, z1, z2) som gir
#like enheter på begge akser. Vi holder x1, x2 og z1 faste og bestemmer z2 slik at (x2-x1)/(z2-z1) i plt=c(x1,x2,z1,z2)
#blir lik grl1/grl3
#Det krever at z2 = [(x2-x1)*grl3 + z1*grl1]/grl1 
#Setter vi x1, x2 og z1 konstante, får vi:
z2 <- (0.8*grl3+0.15*grl1)/grl1
z2

#Diagram for akse 1 mot akse 4
#Vil finne koordinatene for hjørner i det 'plotteområdet' med koordinater (x1, x2, w1, w2) som gir
#like enheter på begge akser. Vi holder x1, x2 og w1 faste og bestemmer w2 slik at (x2-x1)/(w2-w1) i plt=c(x1,x2,w1,w2)
#blir lik grl1/grl4
#Det krever at w2 = [(x2-x1)*grl3 + w1*grl1]/grl1 
#Setter vi x1, x2 og w1 konstante, får vi:
w2 <- (0.8*grl4+0.15*grl1)/grl1
w2
```


```{r, cache = TRUE, include=FALSE}
par(mfrow=c(1,1))

#Tegner diagrammet for aksene 1 og 2
par(plt=c(0.15, 0.95, 0.25, y2+0.25))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
points(akse1,akse2,pch=16,cex=0.75)
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

##Vektorer for nøkkelvariabler, med envfit, diagrammet for akse 1 og 2; merk at r2 er squared Pearson's 

#Setter sammen de aktuelle aksene til en dataramme
ordx <- data.frame(akse1, akse2, akse3, akse4)
is.data.frame(ordx)
dim(ordx)

#Samler aktuelle nøkkelvariabler + 'surrogater' i ei dataramme
nvxOI <- data.frame(Gab_a,IfI,Lled_a,Vei_b_a,Sefr_a,Cul_a_s,Abygg_a)
nvxOI
nvxJP <- data.frame(Arfull_a,JI)
nvxJP
nvxSN <- data.frame(Arlov_a,Arbar_a,Land_a,Arbla_a,R_net_a,Kbf_a,Oyst_i,Ekspve_a)
nvxSN
nvxIP <- data.frame(Inns_s,II)
nvxIP
nvxMP <- data.frame(MI)
nvxMP
nvxX <- data.frame(Sn_torr,Sn_frisk,Sn_flekk,Bohei_a,Araaf_a,Kskred_a)
nvxX
nvx3 <- data.frame(IfI,Arfull_a,Oyst_i,RR1)
nvx3
nvxRR <- data.frame(RR1, Cre_f_a, Cre_f_a, Flat_a, Tpi1h_a, Tpi1l_a, Rug3_m, Guro_t_a)
nvxRR

#Beregner envfit-vektorer
ordvek12OI <- envfit(ordx[,1:2], nvxOI, permutations = 999)
ordvek12JP <- envfit(ordx[,1:2], nvxJP, permutations = 999)
ordvek12SN <- envfit(ordx[,1:2], nvxSN, permutations = 999)
ordvek12RR <- envfit(ordx[,1:2], nvxRR, permutations = 999)

ordvek12OI$vectors$arrows #gir koordinater for vektorpilenes hoder, standardisert til lengde = 1
ordvek12OI$vectors$r #gir r-squared for variablene
ordvek12OI #gir begge deler, skrevet ut i tabell
ordvek12JP #gir begge deler, skrevet ut i tabell
ordvek12SN #gir begge deler, skrevet ut i tabell
ordvek12RR #gir begge deler, skrevet ut i tabell

ordvek13IP <- envfit(ordx[,c(1,3)], nvxIP, permutations = 999)
ordvek13IP #gir begge deler, skrevet ut i tabell
ordvek13MP <- envfit(ordx[,c(1,3)], nvxMP, permutations = 999)
ordvek13MP
ordvek13X <- envfit(ordx[,c(1,3)], nvxX, permutations = 999)
ordvek13X
ordvek133 <- envfit(ordx[,c(1,3)], nvx3, permutations = 999)
ordvek133
```


```{r, cache = TRUE, echo=FALSE}

#Tegner diagrammet for aksene 1 og 2
par(plt=c(0.15, 0.95, 0.25, (y2+0.25)))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
points(akse1,akse2,pch=16,cex=0.75)
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

#Plotter vektorpiler på standardordinasjonsdiagrammene med kommandoene:
plot(ordvek12OI,arrow.mul=.58,col="brown",add=T,cex=0.8)
plot(ordvek12JP,arrow.mul=.58,col="green3",add=T,cex=0.8)
plot(ordvek12SN,arrow.mul=.58,col="blue3",add=T,cex=0.8)

```

```{r, cache = TRUE, echo=FALSE}

#Tegner diagrammet for aksene 1 og 2
par(plt=c(0.15, 0.95, 0.25, (y2+0.25)))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
points(akse1,akse2,pch=16,cex=0.75)
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

#Plotter vektorpiler på standardordinasjonsdiagrammene med kommandoene:
plot(ordvek12RR,arrow.mul=.49,col="darkviolet",add=T,cex=0.8)
plot(ordvek133,arrow.mul=.61,col="black",add=T,cex=0.6)

```

#### Complementary observations, axis 1 and 2:

* One OU belongs to OI Class 6 (big city). This is Id 2916 Oslo city, which occupies the extreme position along GNMDS axis 1.
* Agricultural intensity is associated with positive scores for GNMDS 1, and a certain degree of infrastructure (similar to IfI ≈ 4) is a prerequisite for amount/degree of agricultural land use intensity. The relationship between JI and IfI is slightly unimodal, that is, the agricultural land use intensity is strongest when the amount/degree of infrastructure is relatively high, but decreases again when the amount/degree of infrastructure is very high.
* The proportion of the area with deciduous forest (Arlov_a) is 0 at the outer parts of the archipelago (SN = 5; there are 156 OUs with Arlov = 0), but increases inward towards larger islands and on the mainland.
* The figure at the bottom left shows a clear (inverse) island size gradient, with a considerable variation in landscape characteristics related to the archipelago properties, first and foremost when the islands are smaller than 100 km2 (Oyst_i > 0.643). This opens the possibility that the complex gradient expressed on axis 1 is actually composed of two CLGs.
* Among the 69 OUs in KS lacking infrastructure (IfI = 0), all of the 56 as a largest island smaller than 1 km. Islands with significant infrastructure are generally much larger; one exception is Id 4038 which has Oyst_i = 0.8647 and IfI = 9.34. This polygon is located on the west and south part of Nøtterøy.
* It is noteworthy that the topographic variables give such a strong signal when there is so little topography variation within costal plains.

```{r, fig.height = 4.7, fig.width = 10.5,echo=FALSE}
###Figurer med isolinjer på ordinasjonsdiagrammet (Ordisurf); akser 1 mot 2.
par(mfrow=c(1,2))

##For IfI med OI (akser 1 og 2)

par(plt=c(0.15, 0.95, 0.25, (y2+0.25)))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

os <- ordisurf(ordx,IfI,display="sites",levels=c(1,2,4,6,8,10,11,12,13,14), col="red",add=T) #inkluderer plotting på eksisterende ordinasjonsdiagram
lines(c(-0.77,-0.74),c(0.535,0.535),col=2)
text(-0.72,0.535,"IfI",adj=0,cex=0.8)

#Punkter ulike farger og symboler for ulike infrastruktur-trinn (OI-klasser 1-6)
points(akse1[OI==1],akse2[OI==1],pch=1,cex=0.6,col="lightblue3")
points(akse1[OI==2],akse2[OI==2],pch=16,cex=0.6,col="lightblue3")
points(akse1[OI==3],akse2[OI==3],pch=1,cex=0.75,col="lightblue4")
points(akse1[OI==4],akse2[OI==4],pch=16,cex=0.75,col="lightblue4")
points(akse1[OI==5],akse2[OI==5],pch=1,cex=0.9,col="red")
points(akse1[OI==6],akse2[OI==6],pch=16,cex=0.9,col="red")
points(-0.75,0.49,pch=1,cex=0.6,col="lightblue3")
text(-0.72,0.49,"OI = 1",adj=0,cex=0.8)
points(-0.75,0.45,pch=16,cex=0.6,col="lightblue3")
text(-0.72,0.45,"OI = 2",adj=0,cex=0.8)
points(-0.75,0.41,pch=1,cex=0.75,col="lightblue4")
text(-0.72,0.41,"OI = 3",adj=0,cex=0.8)
points(-0.75,0.37,pch=16,cex=0.75,col="lightblue4")
text(-0.72,0.37,"OI = 4",adj=0,cex=0.8)
points(-0.75,0.33,pch=1,cex=0.9,col="red")
text(-0.72,0.33,"OI = 5",adj=0,cex=0.8)
points(-0.75,0.29,pch=16,cex=0.9,col="red")
text(-0.72,0.29,"OI = 6",adj=0,cex=0.8)


par(plt=c(0.15, 0.95, 0.25, y2+0.25))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

os <- ordisurf(ordx,JI,display="sites",levels=c(0.01,0.025,0.05,0.1,0.2), col="red",add=T) #inkluderer plotting på eksisterende ordinasjonsdiagram
lines(c(-0.77,-0.74),c(0.535,0.535),col=2)
text(-0.72,0.535,"JI",adj=0,cex=0.8)

#Punkter ulike farger og symboler for ulike jordbrukspreg-trinn (JP 1-3)
points(akse1[JP==1],akse2[JP==1],pch=1,cex=0.6,col="lightblue2")
points(akse1[JP==2],akse2[JP==2],pch=16,cex=0.75,col="lightblue4")
points(akse1[JP==3],akse2[JP==3],pch=16,cex=0.9,col="red")
points(-0.75,0.49,pch=1,cex=0.6,col="lightblue2")
text(-0.72,0.49,"JP = 1",adj=0,cex=0.8)
points(-0.75,0.45,pch=16,cex=0.75,col="lightblue4")
text(-0.72,0.45,"JP = 2",adj=0,cex=0.8)
points(-0.75,0.41,pch=16,cex=0.9,col="red")
text(-0.72,0.41,"JP = 3",adj=0,cex=0.8)
```

```{r, fig.height = 4.7, fig.width = 10.5,echo=FALSE}
par(mfrow=c(1,2))
par(plt=c(0.15, 0.95, 0.25, z2+0.25))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

os <- ordisurf(ordx[,c(1,2)],Arlov_a,display="sites",levels=c(0.1,0.25,0.4,0.6,0.8), col="red",add=T) #inkluderer plotting på eksisterende ordinasjonsdiagram
lines(c(-0.77,-0.74),c(0.535,0.535),col=2)
text(-0.72,0.535,"Arlov_a",adj=0,cex=0.8)

#Punkter ulike farger og symboler for ulike (nye) skjærgårdspreg-trinn (SN-klasser 1-4)

points(akse1[SN==1],akse3[SN==1],pch=16,cex=0.5,col=1)
points(akse1[SN==2],akse3[SN==2],pch=16,cex=0.6,col="lightblue2")
points(akse1[SN==3],akse3[SN==3],pch=16,cex=0.75,col="darkblue")
points(akse1[SN==4],akse3[SN==4],pch=1,cex=0.75,col="red")
points(akse1[SN==5],akse3[SN==5],pch=16,cex=0.9,col="red")
points(-0.75,0.49,pch=16,cex=0.5,col="black")
text(-0.72,0.49,"SN = 1 (mainland)",adj=0,cex=0.8)
points(-0.75,0.45,pch=16,cex=0.6,col="lightblue2")
text(-0.72,0.45,"SN = 2",adj=0,cex=0.8)
points(-0.75,0.41,pch=16,cex=0.75,col="darkblue")
text(-0.72,0.41,"SN = 3",adj=0,cex=0.8)
points(-0.75,0.37,pch=1,cex=0.75,col="red")
text(-0.72,0.37,"SN = 4",adj=0,cex=0.8)
points(-0.75,0.33,pch=16,cex=0.9,col="red")
text(-0.72,0.33,"SN = 5",adj=0,cex=0.8)

##For Oyst_i

par(plt=c(0.15, 0.95, 0.25, z2+0.25))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

os <- ordisurf(ordx[,c(1,2)],Oyst_i,display="sites",levels=c(0.19,0.454,0.574,0.643,0.693,0.733,0.768,0.798), col="red",add=T) #inkluderer plotting på eksisterende ordinasjonsdiagram
lines(c(-0.77,-0.74),c(0.535,0.535),col=2)
text(-0.72,0.535,"Inverse island size (Oyst_i)",adj=0,cex=0.8)

#Punkter ulike farger og symboler for ulike (nye) skjærgårdspreg-trinn (SN-klasser 1-4)

points(akse1[SN==1],akse3[SN==1],pch=16,cex=0.5,col=1)
points(akse1[SN==2],akse3[SN==2],pch=16,cex=0.6,col="lightblue2")
points(akse1[SN==3],akse3[SN==3],pch=16,cex=0.75,col="darkblue")
points(akse1[SN==4],akse3[SN==4],pch=1,cex=0.75,col="red")
points(akse1[SN==5],akse3[SN==5],pch=16,cex=0.9,col="red")
points(-0.75,0.49,pch=16,cex=0.5,col="black")
text(-0.72,0.49,"SN = 1 (mainland)",adj=0,cex=0.8)
points(-0.75,0.45,pch=16,cex=0.6,col="lightblue2")
text(-0.72,0.45,"SN = 2",adj=0,cex=0.8)
points(-0.75,0.41,pch=16,cex=0.75,col="darkblue")
text(-0.72,0.41,"SN = 3",adj=0,cex=0.8)
points(-0.75,0.37,pch=1,cex=0.75,col="red")
text(-0.72,0.37,"SN = 4",adj=0,cex=0.8)
points(-0.75,0.33,pch=16,cex=0.9,col="red")
text(-0.72,0.33,"SN = 5",adj=0,cex=0.8)
```


```{r, fig.height = 4.7, fig.width = 10.5,echo=FALSE}
par(mfrow=c(1,2))
par(plt=c(0.15, 0.95, 0.25, z2+0.25))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

os <- ordisurf(ordx[,c(1,2)],Kbf_a,display="sites",levels=c(0.2,0.4,0.6,0.8,0.9,0.95), col="red",add=T) #inkluderer plotting p? eksisterende 

#ordinasjonsdiagram
lines(c(-0.77,-0.74),c(0.535,0.535),col=2)
text(-0.72,0.535,"Proportion of area with soil type bare rock (Kbf_a)",adj=0,cex=0.8)

#Punkter ulike farger og symboler for ulike (nye) skj?rg?rdspreg-trinn (SN-klasser 1-4)

points(akse1[SN==1],akse3[SN==1],pch=16,cex=0.5,col=1)
points(akse1[SN==2],akse3[SN==2],pch=16,cex=0.6,col="lightblue2")
points(akse1[SN==3],akse3[SN==3],pch=16,cex=0.75,col="darkblue")
points(akse1[SN==4],akse3[SN==4],pch=1,cex=0.75,col="red")
points(akse1[SN==5],akse3[SN==5],pch=16,cex=0.9,col="red")
points(-0.75,0.49,pch=16,cex=0.5,col="black")
text(-0.72,0.49,"SN = 1 (mainland)",adj=0,cex=0.8)
points(-0.75,0.45,pch=16,cex=0.6,col="lightblue2")
text(-0.72,0.45,"SN = 2",adj=0,cex=0.8)
points(-0.75,0.41,pch=16,cex=0.75,col="darkblue")
text(-0.72,0.41,"SN = 3",adj=0,cex=0.8)
points(-0.75,0.37,pch=1,cex=0.75,col="red")
text(-0.72,0.37,"SN = 4",adj=0,cex=0.8)
points(-0.75,0.33,pch=16,cex=0.9,col="red")
text(-0.72,0.33,"SN = 5",adj=0,cex=0.8)

par(plt=c(0.15, 0.95, 0.25, z2+0.25))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

os <- ordisurf(ordx[,c(1,2)],RR1,display="sites",levels=c(25,50,75,100,150,200), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
lines(c(-0.77,-0.74),c(0.535,0.535),col=2)
text(-0.72,0.535,"RR1",adj=0,cex=0.8)

#Punkter ulike farger og symboler for ulike (nye) skj?rg?rdspreg-trinn (SN-klasser 1-4)

points(akse1[SN==1],akse3[SN==1],pch=16,cex=0.5,col=1)
points(akse1[SN==2],akse3[SN==2],pch=16,cex=0.6,col="lightblue2")
points(akse1[SN==3],akse3[SN==3],pch=16,cex=0.75,col="darkblue")
points(akse1[SN==4],akse3[SN==4],pch=1,cex=0.75,col="red")
points(akse1[SN==5],akse3[SN==5],pch=16,cex=0.9,col="red")
points(-0.75,0.49,pch=16,cex=0.5,col="black")
text(-0.72,0.49,"SN = 1 (mainland)",adj=0,cex=0.8)
points(-0.75,0.45,pch=16,cex=0.6,col="lightblue2")
text(-0.72,0.45,"SN = 2",adj=0,cex=0.8)
points(-0.75,0.41,pch=16,cex=0.75,col="darkblue")
text(-0.72,0.41,"SN = 3",adj=0,cex=0.8)
points(-0.75,0.37,pch=1,cex=0.75,col="red")
text(-0.72,0.37,"SN = 4",adj=0,cex=0.8)
points(-0.75,0.33,pch=16,cex=0.9,col="red")
text(-0.72,0.33,"SN = 5",adj=0,cex=0.8)

```







```{r, fig.height = 4.7, fig.width = 10.5,echo=FALSE}
par(mfrow=c(1,2))

#Tegner diagrammet for aksene 1 og 2
par(plt=c(0.15, 0.95, 0.25, (y2+0.25)))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
points(akse1,akse2,pch=16,cex=0.75)
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

#Plotter vektorpiler på standardordinasjonsdiagrammene med kommandoene:
#plot(ordvek12OI,arrow.mul=.58,col="brown",add=T,cex=0.8)
#plot(ordvek12JP,arrow.mul=.58,col="green3",add=T,cex=0.8)
#plot(ordvek12OI,arrow.mul=.58,col="brown",add=T,cex=0.8)
#plot(ordvek12JP,arrow.mul=.58,col="green3",add=T,cex=0.8)
#plot(ordvek12SN,arrow.mul=.58,col="blue3",add=T,cex=0.8)
#plot(ordvek12RR,arrow.mul=.49,col="darkviolet",add=T,cex=0.8)
#
plot(ordvek13IP,arrow.mul=.73,col="blue3",add=T,cex=0.6)
plot(ordvek13MP,arrow.mul=1.02,col="green3",add=T,cex=0.6)
plot(ordvek13X,arrow.mul=.61,col="brown",add=T,cex=0.6)
plot(ordvek133,arrow.mul=.61,col="black",add=T,cex=0.6)

#par(plt=c(0.15, 0.95, 0.15, z2))
#plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains) (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.


```

Investigated whether there could be grounds for dividing KS into two major types based on dominant soil, parallel to the classification of inland irregular plains by looking at the variation in occurrence of selected soil classes JK1 (based on 50% proportion of area) and JK2 (based of 25% proportion of area) in the ordination chart.
For JK1 the distribution is: B (> 50% bare rock): 346; H (> 50% marine deposits): 27; 0 (none) 61. For JK2 (> 25% proportion of area), 83 OUs belong to H.
The distribution of the earth classes in the GNMDS diagram with axis 1 and 2 shows that there is a relationship between earth and axis 1, but that it is not very strong (there are scattered OUs with bare rock scattered all the way between OUs with larger or less amount of marine deposits).


```{r, echo=FALSE}
##Fig. med jordartskategorier på akser 1 og 2

par(plt=c(0.15, 0.95, 0.25, (z2+0.25)))
plot(akse1,akse2,xlab="GNMDS 1 Coastal plains (KS457)",ylab="GNMDS 2 Coastal plains (KS457)",xlim=c(min(akse1)-0.01,max(akse1)+0.01),ylim=c(min(akse2)-0.01,max(akse2)+0.01),xaxp=c(-0.5,1,3),yaxp=c(-0.5,0.5,2),type="n") #Grunnlag for plotting av symboler etc.
lines(c(-2,2),c(0,0),lty=2,col=8)
lines(c(0,0),c(-2,2),lty=2,col=8)

points(akse1[JK2=="H"],akse2[JK2=="H"],pch=16,cex=0.5,col="red")
points(akse1[JK1=="H"],akse2[JK1=="H"],pch=16,cex=0.8,col="red3")
points(akse1[JK1=="B"],akse2[JK1=="B"],pch=16,cex=0.4,col="black")
points(akse1[JK1==0],akse2[JK1==0],pch=1,cex=0.5,col="lightblue4")

points(-0.75,0.41,pch=1,cex=0.5,col="lightblue4")
text(-0.72,0.41,"JK1 = 0 (ingen)",adj=0,cex=0.8)
points(-0.75,0.45,pch=16,cex=0.4,col="black")
text(-0.72,0.45,"JK1 = B (> 50 % NB)",adj=0,cex=0.8)
points(-0.75,0.49,pch=16,cex=0.5,col="red")
text(-0.72,0.49,"JK2 = H (> 25 % MA)",adj=0,cex=0.8)
points(-0.75,0.53,pch=16,cex=0.8,col="red3")
text(-0.72,0.53,"JK2 = H (> 50 % MA)",adj=0,cex=0.8)
```

Conclusion:

There is no basis for dividing coastal plains into two major types on the basis of dominant soil.

Parallel RDA analyzes with factor variables one and the continuous analysis variables as alternative ways of representing soil variation showed that both single analysis variables Kmar_a and Kbf_a explained more variation in the data set KS457 (0.06004 and 0.05969) than factor variables JK1 and JK2 (0.0387, respectively). Therefore, used the analysis variables to characterize soil as a CLG candidate.
The ordination analysis and interpretation identified the following hCLG candidates by the three functional variables categories G, U and A (thematically slightly marginal variables are highlighted in gray; the absolute value of the correlation coefficient with the relevant GNMDS axis in parentheses):

(G1) Relieff (RE; Axis 2): Crug9_m (0.7286) RR1 (0.7185), Crug3_m (0.6764), Rug3_m (0.6617), Steep_a (0.6572), Cr3_u_a (0.5767), Flat_a (0.5626), Tpi1l_a (0.5585), Guro_t_a (0.5437), Cre_f_a (0.5412), Cre_b_a (0.5234), Tpi1h_a (0.4750) 
(G2) Indre-outer coast (IYK; Axis 1/2): Land_a (0.5702), Stromn_a (0.5203), Oyst_i (0.5057), R_net_a (0.5440), Ekspve_a (0.4351) 
(G3) Vannpreg (VP; Axis 1/3/4: Inns_s (0.4005), Dismire (0.3505), MI (0.3310), II (0.3051), Dislake (0.3049)


#############################################################################################
#### Tallfesting av KLG-kandidatenes relative viktighet ved variasjonsoppdelingsanalyse ####
#############################################################################################



```{r, echo=TRUE}
#Datasett: av

#Bestemmer total variasjon

vo0 <- rda(av~+1,scale=T)  #scale = T gjør at dataene blir sentrert og standardisert ved deling med SD
vo0
```


```{r, echo=TRUE}
voG1a <- rda(av~RR1,scale=T)
voG1a
```


```{r, echo=TRUE}
tvoG1a <- permutest(voG1a,permutations=9999)
tvoG1a
```

```{r}
voG1b <- rda(av~Crug9_m,scale=T)
voG1b
tvoG1b <- permutest(voG1b,permutations=9999)
tvoG1b
```


voG1c <- rda(av~Rug3_m,scale=T)
voG1c
tvoG1c <- permutest(voG1c,permutations=9999)
tvoG1c
voG1d <- rda(av~Cr3_u_a,scale=T)
voG1d
tvoG1d <- permutest(voG1d,permutations=9999)
tvoG1d
voG1e <- rda(av~Flat_a,scale=T)
voG1e
tvoG1e <- permutest(voG1e,permutations=9999)
tvoG1e
voG1f <- rda(av~Tpi1l_a,scale=T)
voG1f
tvoG1f <- permutest(voG1f,permutations=9999)
tvoG1f
voG1g <- rda(av~Guro_t_a,scale=T)
voG1g
tvoG1g <- permutest(voG1g,permutations=9999)
tvoG1g
voG1h <- rda(av~Cre_f_a,scale=T)
voG1h
tvoG1h <- permutest(voG1h,permutations=9999)
tvoG1h
voG1i <- rda(av~Cre_b_a,scale=T)
voG1i
tvoG1i <- permutest(voG1i,permutations=9999)
tvoG1i
voG1j <- rda(av~Tpi1h_a,scale=T)
voG1j
tvoG1j <- permutest(voG1j,permutations=9999)
tvoG1j
voG1k <- rda(av~Steep_a,scale=T)
voG1k
tvoG1k <- permutest(voG1k,permutations=9999)
tvoG1k
voG1l <- rda(av~Crug3_m,scale=T)
voG1l
tvoG1l <- permutest(voG1l,permutations=9999)
tvoG1l
```{r}
voG1ca <- rda(av~RR1+Condition(Rug3_m),scale=T)
voG1ca
tvoG1ca <- permutest(voG1ca,permutations=9999)
tvoG1ca
```

voG1cb <- rda(av~Crug9_m+Condition(Rug3_m),scale=T)
voG1cb
tvoG1cb <- permutest(voG1cb,permutations=9999)
tvoG1cb
```{r}

voG1cd <- rda(av~Cr3_u_a+Condition(Rug3_m),scale=T)
voG1cd
tvoG1cd <- permutest(voG1cd,permutations=9999)
tvoG1cd

```
voG1ce <- rda(av~Flat_a+Condition(Rug3_m),scale=T)
voG1ce
tvoG1ce <- permutest(voG1ce,permutations=9999)
tvoG1ce
voG1cf <- rda(av~Tpi1l_a+Condition(Rug3_m),scale=T)
voG1cf
tvoG1cf <- permutest(voG1cf,permutations=9999)
tvoG1cf
voG1cg <- rda(av~Guro_t_a+Condition(Rug3_m),scale=T)
voG1cg
tvoG1cg <- permutest(voG1cg,permutations=9999)
tvoG1cg
voG1ch <- rda(av~Cre_f_a+Condition(Rug3_m),scale=T)
voG1ch
tvoG1ch <- permutest(voG1ch,permutations=9999)
tvoG1ch
voG1ci <- rda(av~Cre_b_a+Condition(Rug3_m),scale=T)
voG1ci
tvoG1ci <- permutest(voG1ci,permutations=9999)
tvoG1ci
voG1cj <- rda(av~Tpi1h_a+Condition(Rug3_m),scale=T)
voG1cj
tvoG1cj <- permutest(voG1cj,permutatioks=9999)
tvoG1cj
voG1ck <- rda(av~Steep_a+Condition(Rug3_m),scale=T)
voG1ck
tvoG1ck <- permutest(voG1ck,permutations=9999)
tvoG1ck
voG1cl <- rda(av~Crug3_m+Condition(Rug3_m),scale=T)
voG1cl
tvoG1cl <- permutest(voG1cl,permutations=9999)
tvoG1cl

voG1cja <- rda(av~RR1+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cja
tvoG1cja <- permutest(voG1cja,permutations=9999)
tvoG1cja

voG1cjb <- rda(av~Crug9_m+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cjb
tvoG1cjb <- permutest(voG1cjb,permutations=9999)
tvoG1cjb
voG1cjd <- rda(av~Cr3_u_a+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cjd
tvoG1cjd <- permutest(voG1cjd,permutations=9999)
tvoG1cjd
voG1cje <- rda(av~Flat_a+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cje
tvoG1cje <- permutest(voG1cje,permutations=9999)
tvoG1cje
voG1cjf <- rda(av~Tpi1l_a+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cjf
tvoG1cjf <- permutest(voG1cjf,permutations=9999)
tvoG1cjf
voG1cjg <- rda(av~Guro_t_a+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cjg
tvoG1cjg <- permutest(voG1cjg,permutations=9999)
tvoG1cjg
voG1cjh <- rda(av~Cre_f_a+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cjh
tvoG1cjh <- permutest(voG1cjh,permutations=9999)
tvoG1cjh
voG1cji <- rda(av~Cre_b_a+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cji
tvoG1cji <- permutest(voG1cji,permutations=9999)
tvoG1cji
voG1cjl <- rda(av~Crug3_m+Condition(Rug3_m+Tpi1h_a),scale=T)
voG1cjl
tvoG1cjl <- permutest(voG1cjl,permutations=9999)
tvoG1cjl

voG1cjhf <- rda(av~Tpi1l_a+Condition(Rug3_m+Tpi1h_a+Cre_f_a),scale=T)
voG1cjhf
tvoG1cjhf <- permutest(voG1cjhf,permutations=9999)
tvoG1cjhf
voG1cjhk <- rda(av~Crug3_m+Condition(Rug3_m+Tpi1h_a+Cre_f_a),scale=T)
voG1cjhk
tvoG1cjhk <- permutest(voG1cjhk,permutations=99)
tvoG1cjhk

voG2a <- rda(av~Oyst_i,scale=T)
voG2a
tvoG2a <- permutest(voG2a,permutations=9999)
tvoG2a
voG2b <- rda(av~Land_a,scale=T)
voG2b
tvoG2b <- permutest(voG2b,permutations=9999)
tvoG2b
voG2c <- rda(av~R_net_a,scale=T)
voG2c
tvoG2c <- permutest(voG2c,permutations=9999)
tvoG2c
voG2d <- rda(av~Kbf_a,scale=T)
voG2d
tvoG2d <- permutest(voG2d,permutations=9999)
tvoG2d
voG2e <- rda(av~Ekspve_a,scale=T)
voG2e
tvoG2e <- permutest(voG2e,permutations=9999)
tvoG2e
voG2f <- rda(av~Stromn_a,scale=T)
voG2f
tvoG2f <- permutest(voG2f,permutations=9999)
tvoG2f

voG2ca <- rda(av~Oyst_i+Condition(R_net_a),scale=T)
voG2ca
tvoG2ca <- permutest(voG2ca,permutations=99)
tvoG2ca
voG2cb <- rda(av~Land_a+Condition(R_net_a),scale=T)
voG2cb
tvoG2cb <- permutest(voG2cb,permutations=99)
tvoG2cb
voG2cd <- rda(av~Kbf_a+Condition(R_net_a),scale=T)
voG2cd
tvoG2cd <- permutest(voG2cd,permutations=99)
tvoG2cd
voG2ce <- rda(av~Ekspve_a+Condition(R_net_a),scale=T)
voG2ce
tvoG2ce <- permutest(voG2ce,permutations=99)
tvoG2ce
voG2cf <- rda(av~Stromn_a+Condition(R_net_a),scale=T)
voG2cf
tvoG2cf <- permutest(voG2cf,permutations=99)
tvoG2cf

voG2cea <- rda(av~Oyst_i+Condition(R_net_a+Ekspve_a),scale=T)
voG2cea
tvoG2cea <- permutest(voG2cea,permutations=99)
tvoG2cea
voG2ced <- rda(av~Kbf_a+Condition(R_net_a+Ekspve_a),scale=T)
voG2ced
tvoG2ced <- permutest(voG2ced,permutations=99)
tvoG2ced
voG2cef <- rda(av~Stromn_a+Condition(R_net_a+Ekspve_a),scale=T)
voG2cef
tvoG2cef <- permutest(voG2cef,permutations=99)
tvoG2cef

voG2cead <- rda(av~Kbf_a+Condition(R_net_a+Ekspve_a+Oyst_i),scale=T)
voG2cead
tvoG2cead <- permutest(voG2cead,permutations=99)
tvoG2cead
voG2ceaf <- rda(av~Stromn_a+Condition(R_net_a+Ekspve_a+Oyst_i),scale=T)
voG2ceaf
tvoG2ceaf <- permutest(voG2ceaf,permutations=99)
tvoG2ceaf

voG3a <- rda(av~Inns_s,scale=T)
voG3a
tvoG3a <- permutest(voG3a,permutations=9999)
tvoG3a
voG3b <- rda(av~II,scale=T)
voG3b
tvoG3b <- permutest(voG3b,permutations=9999)
tvoG3b
voG3c <- rda(av~Dislake,scale=T)
voG3c
tvoG3c <- permutest(voG3c,permutations=9999)
tvoG3c
voG3d <- rda(av~MI,scale=T)
voG3d
tvoG3d <- permutest(voG3d,permutations=9999)
tvoG3d
voG3e <- rda(av~Dismire,scale=T)
voG3e
tvoG3e <- permutest(voG3e,permutations=9999)
tvoG3e

voG3ea <- rda(av~Inns_s+Condition(Dismire),scale=T)
voG3ea
tvoG3ea <- permutest(voG3ea,permutations=99)
tvoG3ea
voG3eb <- rda(av~II+Condition(Dismire),scale=T)
voG3eb
tvoG3eb <- permutest(voG3eb,permutations=99)
tvoG3eb
voG3ec <- rda(av~Dislake+Condition(Dismire),scale=T)
voG3ec
tvoG3ec <- permutest(voG3ec,permutations=99)
tvoG3ec
voG3ed <- rda(av~MI+Condition(Dismire),scale=T)
voG3ed
tvoG3ed <- permutest(voG3ed,permutations=99)
tvoG3ed

voG3eac <- rda(av~Dislake+Condition(Dismire+Inns_s),scale=T)
voG3eac
tvoG3eac <- permutest(voG3eac,permutations=99)
tvoG3eac

voG4a <- rda(av~C_kk_a,scale=T)
voG4a
tvoG4a <- permutest(voG4a,permutations=99)
tvoG4a
voG4b <- rda(av~Ccom_m,scale=T)
voG4b
tvoG4b <- permutest(voG4b,permutations=99)
tvoG4b
voG4c <- rda(av~Maro_s,scale=T)
voG4c
tvoG4c <- permutest(voG4c,permutations=99)
tvoG4c
voG4d <- rda(av~Bplu_a,scale=T)
voG4d
tvoG4d <- permutest(voG4d,permutations=99)
tvoG4d
voG4e <- rda(av~Kmar_a,scale=T)
voG4e
tvoG4e <- permutest(voG4e,permutations=99)
tvoG4e

voG4ba <- rda(av~C_kk_a+Condition(Ccom_m),scale=T)
voG4ba
tvoG4ba <- permutest(voG4ba,permutations=99)
tvoG4ba
voG4bc <- rda(av~Maro_s+Condition(Ccom_m),scale=T)
voG4bc
tvoG4bc <- permutest(voG4bc,permutations=99)
tvoG4bc
voG4bd <- rda(av~Bplu_a+Condition(Ccom_m),scale=T)
voG4bd
tvoG4bd <- permutest(voG4bd,permutations=99)
tvoG4bd
voG4be <- rda(av~Kmar_a+Condition(Ccom_m),scale=T)
voG4be
tvoG4be <- permutest(voG4be,permutations=99)
tvoG4be

voG4bca <- rda(av~C_kk_a+Condition(Ccom_m+Maro_s),scale=T)
voG4bca
tvoG4bca <- permutest(voG4bca,permutations=99)
tvoG4bca
voG4bcd <- rda(av~Bplu_a+Condition(Ccom_m+Maro_s),scale=T)
voG4bcd
tvoG4bcd <- permutest(voG4bcd,permutations=99)
tvoG4bcd

voG5 <- rda(av~JK1,scale=T)
voG5
tvoG5 <- permutest(voG5,permutations=99)
tvoG5

voG6 <- rda(av~JK2,scale=T)
voG6
tvoG6 <- permutest(voG6,permutations=99)
tvoG6

voG7a <- rda(av~Kbf_a,scale=T)
voG7a
tvoG7a <- permutest(voG7a,permutations=99)
tvoG7a
voG7b <- rda(av~Kmar_a,scale=T)
voG7b
tvoG7b <- permutest(voG7b,permutations=99)
tvoG7b
voG7ba <- rda(av~Kbf_a+Condition(Kmar_a),scale=T)
voG7ba
tvoG7ba <- permutest(voG7ba,permutations=99)
tvoG7ba

voU1a <- rda(av~Arlov_a,scale=T)
voU1a
tvoU1a <- permutest(voU1a,permutations=99)
tvoU1a
voU1b <- rda(av~Arbar_a,scale=T)
voU1b
tvoU1b <- permutest(voU1b,permutations=99)
tvoU1b
voU1c <- rda(av~Arbla_a,scale=T)
voU1c
tvoU1c <- permutest(voU1c,permutations=99)
tvoU1c

voU1ab <- rda(av~Arbar_a+Condition(Arlov_a),scale=T)
voU1ab
tvoU1ab <- permutest(voU1ab,permutations=99)
tvoU1ab
voU1ac <- rda(av~Arbla_a+Condition(Arlov_a),scale=T)
voU1ac
tvoU1ac <- permutest(voU1ac,permutations=99)
tvoU1ac

voU1acb <- rda(av~Arbar_a+Condition(Arlov_a+Arbla_a),scale=T)
voU1acb
tvoU1acb <- permutest(voU1acb,permutations=99)
tvoU1acb


voU2a <- rda(av~Sn_torr,scale=T)
voU2a
tvoU2a <- permutest(voU2a,permutations=99)
tvoU2a
voU2b <- rda(av~Sn_frisk,scale=T)
voU2b
tvoU2b <- permutest(voU2b,permutations=99)
tvoU2b
voU2c <- rda(av~Sn_flekk,scale=T)
voU2c
tvoU2c <- permutest(voU2c,permutations=99)
tvoU2c
voU2d <- rda(av~Bohei_a,scale=T)
voU2d
tvoU2d <- permutest(voU2d,permutations=99)
tvoU2d
voU2e <- rda(av~Araaf_a,scale=T)
voU2e
tvoU2e <- permutest(voU2e,permutations=99)
tvoU2e

voU2da <- rda(av~Sn_torr+Condition(Bohei_a),scale=T)
voU2da
tvoU2da <- permutest(voU2da,permutations=99)
tvoU2da
voU2db <- rda(av~Sn_frisk+Condition(Bohei_a),scale=T)
voU2db
tvoU2db <- permutest(voU2db,permutations=99)
tvoU2db
voU2dc <- rda(av~Sn_flekk+Condition(Bohei_a),scale=T)
voU2dc
tvoU2dc <- permutest(voU2dc,permutations=99)
tvoU2dc
voU2de <- rda(av~Araaf_a+Condition(Bohei_a),scale=T)
voU2de
tvoU2de <- permutest(voU2de,permutations=99)
tvoU2de

voU2dca <- rda(av~Sn_torr+Condition(Bohei_a+Sn_flekk),scale=T)
voU2dca
tvoU2dca <- permutest(voU2dca,permutations=99)
tvoU2dca
voU2dcb <- rda(av~Sn_frisk+Condition(Bohei_a+Sn_flekk),scale=T)
voU2dcb
tvoU2dcb <- permutest(voU2dcb,permutations=99)
tvoU2dcb
voU2dce <- rda(av~Araaf_a+Condition(Bohei_a+Sn_flekk),scale=T)
voU2dce
tvoU2dce <- permutest(voU2dce,permutations=99)
tvoU2dce


voA1a <- rda(av~IfI,scale=T)
voA1a
tvoA1a <- permutest(voA1a,permutations=9999)
tvoA1a
voA1b <- rda(av~Gab_a,scale=T)
voA1b
tvoA1b <- permutest(voA1b,permutations=9999)
tvoA1b
voA1c <- rda(av~Lled_a,scale=T)
voA1c
tvoA1c <- permutest(voA1c,permutations=9999)
tvoA1c
voA1d <- rda(av~Vei_b_a,scale=T)
voA1d
tvoA1d <- permutest(voA1d,permutations=9999)
tvoA1d
voA1e <- rda(av~Sefr_a,scale=T)
voA1e
tvoA1e <- permutest(voA1e,permutations=9999)
tvoA1e
voA1f <- rda(av~Cul_a_s,scale=T)
voA1f
tvoA1f <- permutest(voA1f,permutations=9999)
tvoA1f
voA1g <- rda(av~Abygg_a,scale=T)
voA1g
tvoA1g <- permutest(voA1g,permutations=9999)
tvoA1g
voA1h <- rda(av~Gab_nae,scale=T)
voA1h
tvoA1h <- permutest(voA1h,permutations=99)
tvoA1h

voA1ba <- rda(av~IfI+Condition(Gab_a),scale=T)
voA1ba
tvoA1ba <- permutest(voA1ba,permutations=9999)
tvoA1ba
voA1bc <- rda(av~Lled_a+Condition(Gab_a),scale=T)
voA1bc
tvoA1bc <- permutest(voA1bc,permutations=9999)
tvoA1bc
voA1bd <- rda(av~Vei_b_a+Condition(Gab_a),scale=T)
voA1bd
tvoA1bd <- permutest(voA1bd,permutations=9999)
tvoA1bd
voA1be <- rda(av~Sefr_a+Condition(Gab_a),scale=T)
voA1be
tvoA1be <- permutest(voA1be,permutations=9999)
tvoA1be
voA1bf <- rda(av~Cul_a_s+Condition(Gab_a),scale=T)
voA1bf
tvoA1bf <- permutest(voA1bf,permutations=9999)
tvoA1bf
voA1bg <- rda(av~Abygg_a+Condition(Gab_a),scale=T)
voA1bg
tvoA1bg <- permutest(voA1bg,permutations=9999)
tvoA1bg
voA1bh <- rda(av~Gab_nae+Condition(Gab_a),scale=T)
voA1bh
tvoA1bh <- permutest(voA1bh,permutations=9999)
tvoA1bh

voA1bag <- rda(av~Abygg_a+Condition(Gab_a+IfI),scale=T)
voA1bag
tvoA1bag <- permutest(voA1bag,permutations=9999)
tvoA1bag
voA1bah <- rda(av~Gab_nae+Condition(Gab_a+IfI),scale=T)
voA1bah
tvoA1bah <- permutest(voA1bah,permutations=9999)
tvoA1bah

voA2a <- rda(av~Arfull_a,scale=T)
voA2a
tvoA2a <- permutest(voA2a,permutations=99)
tvoA2a
voA2b <- rda(av~JI,scale=T)
voA2b
tvoA2b <- permutest(voA2c,permutations=99)
tvoA2b
voA2c <- rda(av~Arover_a,scale=T)
voA2c
tvoA2c <- permutest(voA2c,permutations=99)
tvoA2c

voA2ab <- rda(av~JI+Condition(Arfull_a),scale=T)
voA2ab
tvoA2ab <- permutest(voA2ab,permutations=99)
tvoA2ab
voA2ac <- rda(av~Arover_a+Condition(Arfull_a),scale=T)
voA2ac
tvoA2ac <- permutest(voA2ac,permutations=99)
tvoA2ac

voG2G1 <- rda(av~Rug3_m+Tpi1h_a+Condition(R_net_a+Ekspve_a+Oyst_i),scale=T)
voG2G1
tvoG2G1 <- permutest(voG2G1,permutations=99)
tvoG2G1
voG2G3 <- rda(av~Dismire+Inns_s+Dislake+Condition(R_net_a+Ekspve_a+Oyst_i),scale=T)
voG2G3
tvoG2G3 <- permutest(voG2G3,permutations=99)
tvoG2G3
voG2G4 <- rda(av~Ccom_m + Maro_s+Condition(R_net_a+Ekspve_a+Oyst_i),scale=T)
voG2G4
tvoG2G4 <- permutest(voG2G4,permutations=99)
tvoG2G4
voG2G5 <- rda(av~Kbf_a+Kmar_a+Condition(R_net_a+Ekspve_a+Oyst_i),scale=T)
voG2G5
tvoG2G5 <- permutest(voG2G5,permutations=99)
tvoG2G5

voG2G1G3 <- rda(av~Dismire+Inns_s + Dislake+Condition(R_net_a+Ekspve_a+Oyst_i+Rug3_m+Tpi1h_a),scale=T)
voG2G1G3
tvoG2G1G3 <- permutest(voG2G1G3,permutations=99)
tvoG2G1G3
voG2G1G4 <- rda(av~Ccom_m + Maro_s+Condition(R_net_a+Ekspve_a+Oyst_i+Rug3_m+Tpi1h_a),scale=T)
voG2G1G4
tvoG2G1G4 <- permutest(voG2G1G4,permutations=99)
tvoG2G1G4

voG2G1G3G4 <- rda(av~Ccom_m + Maro_s+Condition(R_net_a+Ekspve_a+Oyst_i+Rug3_m+Tpi1h_a+Dismire+Inns_s + Dislake),scale=T)
voG2G1G3G4
tvoG2G1G3G4 <- permutest(voG2G1G3G4,permutations=99)
tvoG2G1G3G4

voGU1 <- rda(av~Arlov_a+Arbla_a+Condition(R_net_a+Ekspve_a+Oyst_i+Rug3_m+Tpi1h_a+Dismire+Inns_s + Dislake),scale=T)
voGU1
tvoGU1 <- permutest(voGU1,permutations=99)
tvoGU1
voGU2 <- rda(av~Bohei_a+Sn_flekk+Condition(R_net_a+Ekspve_a+Oyst_i+Rug3_m+Tpi1h_a+Dismire+Inns_s + Dislake),scale=T)
voGU2
tvoGU2 <- permutest(voGU2,permutations=99)
tvoGU2

voGU1U2 <- rda(av~Bohei_a+Sn_flekk+Condition(Arlov_a+Arbla_a+R_net_a+Ekspve_a+Oyst_i+Rug3_m+Tpi1h_a+Dismire+Inns_s + Dislake),scale=T)
voGU1U2
tvoGU1U2 <- permutest(voGU1U2,permutations=99)
tvoGU1U2


```{r}
voGUA1 <- rda(av~IfI+Gab_a+Condition(Arlov_a+Arbla_a+R_net_a+Ekspve_a+Oyst_i+Rug3_m+Tpi1h_a+Dismire+Inns_s + Dislake),scale=T)
voGUA1
tvoGUA1 <- permutest(voGUA1,permutations=99)
tvoGUA1
```


voGUA2 <- rda(av~Arfull_a+Arover_a+Condition(Arlov_a+Arbla_a+R_net_a+Ekspve_a+Oyst_i+Rug3_m+Tpi1h_a+Dismire+Inns_s + Dislake),scale=T)
voGUA2
tvoGUA2 <- permutest(voGUA2,permutations=99)
tvoGUA2

####################################################################
#### Finner prim?rn?kkelvariabler for identifiserte KLG ved RDA ####
####################################################################

#Skript for ? finne ortogonale prim?rn?kkelvariabler (rONV'er) i ei n?kkelvariabelgruppe

#Om n?dvendig: importerer relevante rPNV'er

#dta_pnv <- read-table("clipboard",header=T)
#attach(dta_pnv)

#VE for 1 .akse i RDA'er gir VA(PNV)
```{r}
RDA_IYK <- rda(av~R_net_a+Ekspve_a  + Oyst_i,scale=T)

```
RDA_RE <- rda(av~Rug3_m + Tpi1h_a,scale=T)
RDA_VP <- rda(av~Dismire+Inns_s + Dislake,scale=T)
RDA_SkP <- rda(av~Arlov_a + Arbla_a,scale=T)
RDA_OI <- rda(av~IfI+Gab_a,scale=T)

#EV for RDA 1 gir VE for PNV
```{r}
summary(RDA_IYK)$cont
```


summary(RDA_RE)$cont
summary(RDA_VP)$cont
summary(RDA_SkP)$cont
summary(RDA_OI)$cont

```{r}
#Finner oPNV'er

oRDA_IYK <- rda(av~R_net_a+Ekspve_a  + Oyst_i,scale=T)
oRDA_RE <- rda(av~Rug3_m + Tpi1h_a+Condition(R_net_a+Ekspve_a  + Oyst_i),scale=T)
oRDA_VP <- rda(av~Dismire+Inns_s + Dislake+Condition(R_net_a+Ekspve_a  + Oyst_i +Rug3_m + Tpi1h_a),scale=T)
oRDA_SkP <- rda(av~Arlov_a + Arbla_a+Condition(Dismire+Inns_s + Dislake+R_net_a+Ekspve_a  + Oyst_i +Rug3_m + Tpi1h_a),scale=T)
oRDA_OI <- rda(av~IfI+Gab_a+Condition(Arlov_a + Arbla_a+Dismire+Inns_s + Dislake+R_net_a+Ekspve_a  + Oyst_i + Rug3_m + Tpi1h_a),scale=T)
```


```{r}
summary(oRDA_IYK)$cont  #gir egenverdien for RDA-akse 1
```
summary(oRDA_RE)$cont
summary(oRDA_VP)$cont
summary(oRDA_SkP)$cont
summary(oRDA_OI)$cont


```{r}
#Ekstraksjon av ONV; dette er vektorer
ONV_IYK <- scores(oRDA_IYK, display="lc", choices=1, scaling=1)
ONV_RE <- -scores(oRDA_RE, display="lc", choices=1, scaling=1)
ONV_VP <- scores(oRDA_VP, display="lc", choices=1, scaling=1)
ONV_SkP <- scores(oRDA_SkP, display="lc", choices=1, scaling=1)
ONV_OI <- -scores(oRDA_OI, display="lc", choices=1, scaling=1)
```



```{r}
#Sjekker om rONV'en har samme fortegn som variablen(e) som karakteriserer den, og gj?r om n?dvendig ny ekstraksjon av sk?rer med motsatt fortegn
cor.test(ONV_IYK,Lled_a,method="k")
cor.test(ONV_RE,RR1,method="k")
cor.test(ONV_VP,Inns_s,method="k")
cor.test(ONV_SkP,Arlov_a,method="k")
cor.test(ONV_OI,IfIu,method="k")
```

```{r}
#Rangerer ONV'ene

rONV_IYK <- (ONV_IYK-min(ONV_IYK))/(max(ONV_IYK)-min(ONV_IYK))
rONV_RE <- (ONV_RE-min(ONV_RE))/(max(ONV_RE)-min(ONV_RE))
rONV_VP <- (ONV_VP-min(ONV_VP))/(max(ONV_VP)-min(ONV_VP))
rONV_SkP <- (ONV_SkP-min(ONV_SkP))/(max(ONV_SkP)-min(ONV_SkP))
rONV_OI <- (ONV_OI-min(ONV_OI))/(max(ONV_OI)-min(ONV_OI))
```

```{r}
head(rONV_IYK)
```

#Eksporterer rONV-sk?rer til Excel

writeClipboard(as.character(rONV_IYK))
writeClipboard(as.character(rONV_RE))
writeClipboard(as.character(rONV_VP))
writeClipboard(as.character(rONV_SkP))
writeClipboard(as.character(rONV_OI))

##########################################################
#### Beregning av LA mellom KLG-klasser ved bruk av PD####
##########################################################

#Importerer om n?dvendig rPNVer fra Excel:

#rPNV <- read.table("clipboard",header=T)
#attach(rPNV)
#names(rPNV)
```{r}
#Analyse av KLG'er representert ved rangerte n?kkelvariabler

rPNV <- rONV_IYK
```


```{r, fig.height = 3, fig.width = 3}
#Inspeksjon av OE'enes frekvensfordeling langs rPNV'en
hist(rPNV)

```

#Importerer landskapsegenskapsprofiler fra regneark

### DENNE FUNKER IKKE ###lep <- read.table("clipboard",header=F)
### DENNE FUNKER IKKE ###attach(lep)
### DENNE FUNKER IKKE ###names(lep)

### DENNE FUNKER IKKE ###dim(lep)

#Beregner artsulikhet (PD)

### DENNE FUNKER IKKE ###dist <- vegdist(lep,method="bray") 
### DENNE FUNKER IKKE ###dist

### DENNE FUNKER IKKE ####PD-verdiene er s? lave at det ikke er noe behov for ? bruke geodetisk avstand

### DENNE FUNKER IKKE ####Eksporterer til Excel

### DENNE FUNKER IKKE ###writeClipboard(as.character(dist)) #M? importeres som 'character' (jf. Crawley 2007: 77)

### DENNE FUNKER IKKE ####Importerer IA-estimater fra NiN2LAD5/OI-PD2

### DENNE FUNKER IKKE ###IAest.df <- read.table("clipboard",header=T)
### DENNE FUNKER IKKE ###attach(IAest.df)
### DENNE FUNKER IKKE ###names(IAest.df)

### DENNE FUNKER IKKE ###hist(IAest)

### DENNE FUNKER IKKE ###mean(IAest)
### DENNE FUNKER IKKE ###sd(IAest)


####################################################################
#### Finner prim?rn?kkelvariabler for identifiserte KLG ved RDA ####
####################################################################

#Skript for ? finne ortogonale prim?rn?kkelvariabler (rONV'er) i ei n?kkelvariabelgruppe

#Om n?dvendig: importerer relevante rPNV'er

#dta_pnv <- read-table("clipboard",header=T)
#attach(dta_pnv)

#VE for 1 .akse i RDA'er gir VA(PNV)
```{r, include=FALSE}
RDA_RE <- rda(av~RR1+Rug3_m+Tpi1h_a,scale=T) #er lik PNV, for den f?rst utvalgte gruppa
RDA_SkP <- rda(av~Arbar_a + Sn_imp,scale=T)
RDA_OI <- rda(av~IfIu+Gab_a,scale=T)
#EV for RDA 1 gir VE for PNV
RDA_RE
RDA_SkP
RDA_OI
```
```{r}
summary(oRDA_RE)$cont
```



```{r}
oRDA_RE <- rda(av~RR1+Rug3_m+Tpi1h_a,scale=T) #er lik PNV, for den f?rst utvalgte gruppa
oRDA_SkP <- rda(av~Arbar_a + Sn_imp+Condition(RR1+Rug3_m+Tpi1h_a),scale=T)
oRDA_OI <- rda(av~IfIu+Gab_a+Condition(Arbar_a + Sn_imp+RR1+Rug3_m+Tpi1h_a),scale=T)

summary(oRDA_RE)$cont
summary(oRDA_SkP)$cont #gir aksenes egenverdier
summary(oRDA_OI)$cont
```



```{r}
#Ekstraksjon av ONV; dette er vektorer
ONV_RE <- scores(oRDA_RE, display="lc", choices=1, scaling=1)
ONV_SkP <- -scores(oRDA_SkP, display="lc", choices=1, scaling=1)
ONV_OI <- -scores(oRDA_OI, display="lc", choices=1, scaling=1)
```

```{r}
#Sjekker om rONV'en har samme fortegn som variablen(e) som karakteriserer den, og gjlrb om n?dvendig ny ekstraksjon av sk?rer med motsatt fortegn
cor.test(ONV_RE,Rr1_m,method="k")
cor.test(ONV_SkP,Arbar_a,method="k")
cor.test(ONV_OI,IfIu,method="k")
```

#Eksporterer evt sk?rer langs ONV'ene til Excel
writeClipboard(as.character(ONV_RE))
writeClipboard(as.character(ONV_SkP))
writeClipboard(as.character(ONV_OI)) 

```{r}
#Rangerer ONV'ene

rONV_RE <- (ONV_RE-min(ONV_RE))/(max(ONV_RE)-min(ONV_RE)) #Tar vare p? de korrigerte prim?rn?kkelvariablene
rONV_SkP <- (ONV_SkP-min(ONV_SkP))/(max(ONV_SkP)-min(ONV_SkP))
rONV_OI <- (ONV_OI-min(ONV_OI))/(max(ONV_OI)-min(ONV_OI))
```



#Eksporterer rONV-sk?rer til Excel

writeClipboard(as.character(rONV_RE))
writeClipboard(as.character(rONV_SkP))
writeClipboard(as.character(rONV_OI))

##########################################################
#### Beregning av LA mellom KLG-klasser ved bruk av PD####
##########################################################

#Importerer om n?dvendig rPNVer fra Excel:

#rPNV <- read.table("clipboard",header=T)
#attach(rPNV)
#names(rPNV)

#Analyse av KLG'er representert ved rangerte n?kkelvariabler
```{r}

rONV <- rONV_OI

#Inspeksjon av OE'enes frekvensfordeling langs rPNV'en
hist(rONV)
```




#Importerer landskapsegenskapsprofiler fra regneark
### DENNE FUNKER IKKE ### 
### DENNE FUNKER IKKE ### lep <- read.table("clipboard",header=F)
### DENNE FUNKER IKKE ###lep <- as.numeric(lep)
### DENNE FUNKER IKKE ###attach(lep)
### DENNE FUNKER IKKE ###names(lep)

### DENNE FUNKER IKKE ###dim(lep)

### DENNE FUNKER IKKE ####Beregner artsulikhet (PD)

### DENNE FUNKER IKKE ###dist <- vegdist(lep,method="bray") 
### DENNE FUNKER IKKE ###dist

### DENNE FUNKER IKKE ####PD-verdiene er s? lave at det ikke er noe behov for ? bruke geodetisk avstand

### DENNE FUNKER IKKE ####Eksporterer til Excel

### DENNE FUNKER IKKE ###writeClipboard(as.character(dist)) #M? importeres som 'character' (jf. Crawley 2007: 77)

### DENNE FUNKER IKKE ###cor.test(IfIu,rONV_AP)

### DENNE FUNKER IKKE ### SLUTT ###

###################################################################
#### Analyser av samvariasjon mellom reskalerte rONV'er (sONV) ####
###################################################################

#Fig. 1. 1 IyK mot 2 RE; SN som symboler og isolinjer for Oyst_i

```{r}
#sONV'er som skal plottes mot hverandre
s1 <- sONV_IyK
s2 <- sONV_RE
```

#Finner koordinatene for hj?rner i det 'plotteomr?det' med koordinater (x1, x2, y1, y2) som gir
#like enheter p? begge akser. Vi holder x1, x2 og y1 faste og bestemmer y2 slik at (x2-x1)/(y2-y1) i plt=c(x1,x2,y1,y2)
#blir lik max(s1)/max(s2)
#Det krever at y2 = [(x2-x1)*max(s2) + y1*max(s1)]/max(s1) 
#Setter vi x1, x2 og y1 konstante, f?r vi:
```{r}
y2 <- (0.8*max(s2)+0.15*max(s1))/max(s1)
y2
```


```{r}
#Plotter SN-klasser p? figuren, og legger p? isolinjer for Oyst_i
par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 2 RE - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

lines(c(0,max(s1)),c(max(s2)/3,max(s2)/3),lty=2,col=8)
lines(c(0,max(s1)),c(2*max(s2)/3,2*max(s2)/3),lty=3,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)),lty=2,col=8)

lines(c(0,0.142),c(max(s2),max(s2)),col=8)
lines(c(0,0.142),c(max(s2)-0.025,max(s2)-0.025),col=8)
lines(c(0.142,0.142),c(max(s2)-0.025,max(s2)),col=8)
lines(c(0,0),c(max(s2)-0.025,max(s2)),col=8)
points(0.01,max(s2)-0.012,pch=16,cex=0.5,col="black")
points(0.02,max(s2)-0.012,pch=16,cex=0.6,col="lightblue2")
points(0.03,max(s2)-0.012,pch=16,cex=0.75,col="darkblue")
points(0.04,max(s2)-0.012,pch=1,cex=0.75,col="red")
points(0.05,max(s2)-0.012,pch=16,cex=0.9,col="red")

text(0.06,max(s2)-0.012,"SN & Oyst_i",adj=0,cex=0.8)
points(s1[SN==1],s2[SN==1],pch=16,cex=0.5,col=1)
points(s1[SN==2],s2[SN==2],pch=16,cex=0.6,col="lightblue2")
points(s1[SN==3],s2[SN==3],pch=16,cex=0.75,col="darkblue")
points(s1[SN==4],s2[SN==4],pch=1,cex=0.75,col="red")
points(s1[SN==5],s2[SN==5],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,2)],Oyst_i,display="sites",levels=c(0.19,0.454,0.574,0.643,0.693,0.733,0.768,0.798), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
  
  #Plotter OI-klasser p? figuren, og legger p? isolinjer for IfIu
  par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 2 RE - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

lines(c(0,max(s1)),c(max(s2)/3,max(s2)/3),lty=2,col=8)
lines(c(0,max(s1)),c(2*max(s2)/3,2*max(s2)/3),lty=3,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)),lty=2,col=8)

lines(c(0,0.142),c(max(s2),max(s2)),col=8)
lines(c(0,0.142),c(max(s2)-0.025,max(s2)-0.025),col=8)
lines(c(0.142,0.142),c(max(s2)-0.025,max(s2)),col=8)
lines(c(0,0),c(max(s2)-0.025,max(s2)),col=8)
points(0.01,max(s2)-0.012,pch=1,cex=0.6,col="lightblue3")
points(0.02,max(s2)-0.012,pch=16,cex=0.6,col="lightblue3")
points(0.03,max(s2)-0.012,pch=1,cex=0.75,col="lightblue4")
points(0.04,max(s2)-0.012,pch=16,cex=0.75,col="lightblue4")
points(0.05,max(s2)-0.012,pch=1,cex=0.9,col="red")
points(0.06,max(s2)-0.012,pch=16,cex=0.9,col="red")

text(0.07,max(s2)-0.012,"OI & IfIu",adj=0,cex=0.8)
points(s1[OI==1],s2[OI==1],pch=1,cex=0.6,col="lightblue3")
points(s1[OI==2],s2[OI==2],pch=16,cex=0.6,col="lightblue3")
points(s1[OI==3],s2[OI==3],pch=1,cex=0.75,col="lightblue4")
points(s1[OI==4],s2[OI==4],pch=16,cex=0.75,col="lightblue4")
points(s1[OI==5],s2[OI==5],pch=1,cex=0.9,col="red")
points(s1[OI==6],s2[OI==6],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,2)],IfIu,display="sites",levels=c(1,2,4,6,8,10,11), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
#Plotter JP-klasser p? figuren, og legger p? isolinjer for Ar_full
  par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 2 RE - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

lines(c(0,max(s1)),c(max(s2)/3,max(s2)/3),lty=2,col=8)
lines(c(0,max(s1)),c(2*max(s2)/3,2*max(s2)/3),lty=3,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)),lty=2,col=8)

lines(c(0,0.142),c(max(s2),max(s2)),col=8)
lines(c(0,0.142),c(max(s2)-0.025,max(s2)-0.025),col=8)
lines(c(0.142,0.142),c(max(s2)-0.025,max(s2)),col=8)
lines(c(0,0),c(max(s2)-0.025,max(s2)),col=8)
points(0.01,max(s2)-0.012,pch=1,cex=0.6,col="lightblue2")
points(0.02,max(s2)-0.012,pch=16,cex=0.75,col="lightblue4")
points(0.03,max(s2)-0.012,pch=16,cex=0.9,col="red")

text(0.04,max(s2)-0.012,"JI & Arfull_a",adj=0,cex=0.8)

points(s1[JP==1],s2[JP==1],pch=1,cex=0.6,col="lightblue2")
points(s1[JP==2],s2[JP==2],pch=16,cex=0.75,col="lightblue4")
points(s1[JP==3],s2[JP==3],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,2)],Arfull_a,display="sites",levels=c(0.1,0.4,0.6), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
#Plotter MI-klasser p? figuren, og legger p? isolinjer for Mire_a
  par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 2 RE - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

lines(c(0,max(s1)),c(max(s2)/3,max(s2)/3),lty=2,col=8)
lines(c(0,max(s1)),c(2*max(s2)/3,2*max(s2)/3),lty=3,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)),lty=2,col=8)

lines(c(0,0.142),c(max(s2),max(s2)),col=8)
lines(c(0,0.142),c(max(s2)-0.025,max(s2)-0.025),col=8)
lines(c(0.142,0.142),c(max(s2)-0.025,max(s2)),col=8)
lines(c(0,0),c(max(s2)-0.025,max(s2)),col=8)
points(0.01,max(s2)-0.012,pch=1,cex=0.6,col="lightblue4")
points(0.02,max(s2)-0.012,pch=16,cex=0.6,col="red")

text(0.03,max(s2)-0.012,"MP & Mire_a",adj=0,cex=0.8)

points(s1[MP==1],s2[MP==1],pch=16,cex=0.6,col="lightblue4")
points(s1[MP==2],s2[MP==2],pch=16,cex=0.6,col="red")

os <- ordisurf(snv[,c(1,2)],Mire_a,display="sites",levels=c(0.05,0.1,0.2,0.4,0.6,0.8), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
#Plotter RE-klasser p? figuren, og legger p? isolinjer for RR1
par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 2 RE - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

lines(c(0,max(s1)),c(max(s2)/3,max(s2)/3),lty=2,col=8)
lines(c(0,max(s1)),c(2*max(s2)/3,2*max(s2)/3),lty=3,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)),lty=2,col=8)

lines(c(0,0.142),c(max(s2),max(s2)),col=8)
lines(c(0,0.142),c(max(s2)-0.025,max(s2)-0.025),col=8)
lines(c(0.142,0.142),c(max(s2)-0.025,max(s2)),col=8)
lines(c(0,0),c(max(s2)-0.025,max(s2)),col=8)
points(0.01,max(s2)-0.012,pch=1,cex=0.6,col="lightblue4")
points(0.02,max(s2)-0.012,pch=16,cex=0.6,col="red")

points(0.01,max(s2)-0.012,pch=16,cex=0.5,col="black")
points(0.02,max(s2)-0.012,pch=16,cex=0.6,col="lightblue2")
points(0.03,max(s2)-0.012,pch=16,cex=0.75,col="darkblue")
points(0.04,max(s2)-0.012,pch=1,cex=0.75,col="red")
points(0.05,max(s2)-0.012,pch=16,cex=0.9,col="red")

text(0.06,max(s2)-0.012,"SN & RR1",adj=0,cex=0.8)
points(s1[SN==1],s2[SN==1],pch=16,cex=0.5,col=1)
points(s1[SN==2],s2[SN==2],pch=16,cex=0.6,col="lightblue2")
points(s1[SN==3],s2[SN==3],pch=16,cex=0.75,col="darkblue")
points(s1[SN==4],s2[SN==4],pch=1,cex=0.75,col="red")
points(s1[SN==5],s2[SN==5],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,2)],RR1,display="sites",levels=c(25,50,75,100,150,200), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram

###
```


```{r}
#Plotter RE-klasser p? figuren, og legger p? isolinjer for Rug3_m
par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 2 RE - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

lines(c(0,max(s1)),c(max(s2)/3,max(s2)/3),lty=2,col=8)
lines(c(0,max(s1)),c(2*max(s2)/3,2*max(s2)/3),lty=3,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)),lty=2,col=8)

lines(c(0,0.142),c(max(s2),max(s2)),col=8)
lines(c(0,0.142),c(max(s2)-0.025,max(s2)-0.025),col=8)
lines(c(0.142,0.142),c(max(s2)-0.025,max(s2)),col=8)
lines(c(0,0),c(max(s2)-0.025,max(s2)),col=8)
points(0.01,max(s2)-0.012,pch=1,cex=0.6,col="lightblue4")
points(0.02,max(s2)-0.012,pch=16,cex=0.6,col="red")

points(0.01,max(s2)-0.012,pch=16,cex=0.5,col="black")
points(0.02,max(s2)-0.012,pch=16,cex=0.6,col="lightblue2")
points(0.03,max(s2)-0.012,pch=16,cex=0.75,col="darkblue")
points(0.04,max(s2)-0.012,pch=1,cex=0.75,col="red")
points(0.05,max(s2)-0.012,pch=16,cex=0.9,col="red")

text(0.06,max(s2)-0.012,"SN & Rug3_m",adj=0,cex=0.8)
points(s1[SN==1],s2[SN==1],pch=16,cex=0.5,col=1)
points(s1[SN==2],s2[SN==2],pch=16,cex=0.6,col="lightblue2")
points(s1[SN==3],s2[SN==3],pch=16,cex=0.75,col="darkblue")
points(s1[SN==4],s2[SN==4],pch=1,cex=0.75,col="red")
points(s1[SN==5],s2[SN==5],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,2)],Rug3_m,display="sites",levels=c(0.1,0.2,0.4,0.6,0.8,0.9), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
#Fig. 2. 1 IyK mot 3 VP; SN som symboler og isolinjer for Inns_s

#sONV'er som skal plottes mot hverandre
s1 <- sONV_IyK
s2 <- sONV_VP

#Finner koordinatene for hj?rner i det 'plotteomr?det' med koordinater (x1, x2, y1, y2) som gir
#like enheter p? begge akser. Vi holder x1, x2 og y1 faste og bestemmer y2 slik at (x2-x1)/(y2-y1) i plt=c(x1,x2,y1,y2)
#blir lik max(s1)/max(s2)
#Det krever at y2 = [(x2-x1)*max(s2) + y1*max(s1)]/max(s1) 
#Setter vi x1, x2 og y1 konstante, f?r vi:
y2 <- (0.8*max(s2)+0.15*max(s1))/max(s1)
#y2

par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 3 VP - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

polygon(c(4*max(s1)/5,4*max(s1)/5,max(s1),max(s1)),c(max(s2)/2,max(s2)-0.025,max(s2)-0.025,max(s2)/2),col="lightgray")
polygon(c(4*max(s1)/5,4*max(s1)/5,max(s1),max(s1)),c(max(s2)/2,max(s2)-0.025,max(s2)-0.025,max(s2)/2),col="lightgray",density=10) #legger egentlig et skr?tt raster opp?, og fjerner streken rundt

lines(c(0,max(s1)),c(max(s2)/2,max(s2)/2),lty=2,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)

lines(c(max(s1)-0.14,max(s1)),c(max(s2),max(s2)),col=8)
lines(c(max(s1)-0.14,max(s1)),c(max(s2)-0.025,max(s2)-0.025),col=8)
lines(c(max(s1)-0.14,max(s1)-0.14),c(max(s2)-0.025,max(s2)),col=8)
lines(c(max(s1),max(s1)),c(max(s2)-0.025,max(s2)),col=8)

text(max(s1)-0.075,max(s2)-0.012,"SN & Inns_s",adj=0,cex=0.8)
points(max(s1)-0.13,max(s2)-0.012,pch=16,cex=0.5,col="black")
points(max(s1)-0.12,max(s2)-0.012,pch=16,cex=0.6,col="lightblue2")
points(max(s1)-0.11,max(s2)-0.012,pch=16,cex=0.75,col="darkblue")
points(max(s1)-0.10,max(s2)-0.012,pch=1,cex=0.75,col="red")
points(max(s1)-0.09,max(s2)-0.012,pch=16,cex=0.9,col="red")

points(s1[SN==1],s2[SN==1],pch=16,cex=0.5,col=1)
points(s1[SN==2],s2[SN==2],pch=16,cex=0.6,col="lightblue2")
points(s1[SN==3],s2[SN==3],pch=16,cex=0.75,col="darkblue")
points(s1[SN==4],s2[SN==4],pch=1,cex=0.75,col="red")
points(s1[SN==5],s2[SN==5],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,3)],Inns_s,display="sites",levels=c(0.1,0.2,0.3,0.4,0.5,0.6), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
###

#1 IyK mot 3 VP; SN som symboler og isolinjer for Mire_a

par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 3 VP - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

polygon(c(4*max(s1)/5,4*max(s1)/5,max(s1),max(s1)),c(max(s2)/2,max(s2)-0.025,max(s2)-0.025,max(s2)/2),col="lightgray")
polygon(c(4*max(s1)/5,4*max(s1)/5,max(s1),max(s1)),c(max(s2)/2,max(s2)-0.025,max(s2)-0.025,max(s2)/2),col="lightgray",density=10) #legger egentlig et skr?tt raster opp?, og fjerner streken rundt

lines(c(0,max(s1)),c(max(s2)/2,max(s2)/2),lty=2,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)

lines(c(max(s1)-0.14,max(s1)),c(max(s2),max(s2)),col=8)
lines(c(max(s1)-0.14,max(s1)),c(max(s2)-0.025,max(s2)-0.025),col=8)
lines(c(max(s1)-0.14,max(s1)-0.14),c(max(s2)-0.025,max(s2)),col=8)
lines(c(max(s1),max(s1)),c(max(s2)-0.025,max(s2)),col=8)

text(max(s1)-0.075,max(s2)-0.012,"SN & Mire_a",adj=0,cex=0.8)
points(max(s1)-0.13,max(s2)-0.012,pch=16,cex=0.5,col="black")
points(max(s1)-0.12,max(s2)-0.012,pch=16,cex=0.6,col="lightblue2")
points(max(s1)-0.11,max(s2)-0.012,pch=16,cex=0.75,col="darkblue")
points(max(s1)-0.10,max(s2)-0.012,pch=1,cex=0.75,col="red")
points(max(s1)-0.09,max(s2)-0.012,pch=16,cex=0.9,col="red")

points(s1[SN==1],s2[SN==1],pch=16,cex=0.5,col=1)
points(s1[SN==2],s2[SN==2],pch=16,cex=0.6,col="lightblue2")
points(s1[SN==3],s2[SN==3],pch=16,cex=0.75,col="darkblue")
points(s1[SN==4],s2[SN==4],pch=1,cex=0.75,col="red")
points(s1[SN==5],s2[SN==5],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,3)],Mire_a,display="sites",levels=c(0.05,0.1,0.2,0.3,0.4,0.5,0.6), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
#Fig. 3. 2 RE mot 3 VP; SN som symboler og isolinjer for Inns_s
##Ikke relevant for siste gangs analyse

#sONV'er som skal plottes mot hverandre
s1 <- sONV_RE
s2 <- sONV_VP

#Finner koordinatene for hj?rner i det 'plotteomr?det' med koordinater (x1, x2, y1, y2) som gir
#like enheter p? begge akser. Vi holder x1, x2 og y1 faste og bestemmer y2 slik at (x2-x1)/(y2-y1) i plt=c(x1,x2,y1,y2)
#blir lik max(s1)/max(s2)
#Det krever at y2 = [(x2-x1)*max(s2) + y1*max(s1)]/max(s1) 
#Setter vi x1, x2 og y1 konstante, f?r vi:
y2 <- (0.8*max(s2)+0.15*max(s1))/max(s1)
y2

par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 2 RE - Coastal plains (KS457)",ylab="KLG 3 VP - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

polygon(c(max(s1)/2,max(s1),max(s1),max(s1)/2),c(max(s2)/2,max(s2)/2,max(s2),max(s2)),col="lightgray")
polygon(c(max(s1)/2,max(s1),max(s1),max(s1)/2),c(max(s2)/2,max(s2)/2,max(s2),max(s2)),col="lightgray",density=10) #legger egentlig et skr?tt raster opp?, og fjerner streken rundt

lines(c(0,max(s1)),c(max(s2)/2,max(s2)/2),lty=2,col=8)
lines(c(max(s1)/2,max(s1)/2),c(0,max(s2)),lty=2,col=8)

lines(c(0,0.095),c(max(s2),max(s2)),col=8)
lines(c(0,0.095),c(max(s2)-0.02,max(s2)-0.02),col=8)
lines(c(0.095,0.095),c(max(s2)-0.02,max(s2)),col=8)
lines(c(0,0),c(max(s2)-0.02,max(s2)),col=8)

points(0.007,max(s2)-0.01,pch=16,cex=0.5,col="black")
points(0.014,max(s2)-0.01,pch=16,cex=0.6,col="lightblue2")
points(0.021,max(s2)-0.01,pch=16,cex=0.75,col="darkblue")
points(0.028,max(s2)-0.01,pch=1,cex=0.75,col="red")
points(0.035,max(s2)-0.01,pch=16,cex=0.9,col="red")

text(0.045,max(s2)-0.01,"SN & Inns_s",adj=0,cex=0.8)
points(s1[SN==1],s2[SN==1],pch=16,cex=0.5,col=1)
points(s1[SN==2],s2[SN==2],pch=16,cex=0.6,col="lightblue2")
points(s1[SN==3],s2[SN==3],pch=16,cex=0.75,col="darkblue")
points(s1[SN==4],s2[SN==4],pch=1,cex=0.75,col="red")
points(s1[SN==5],s2[SN==5],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,2)],Inns_s,display="sites",levels=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
###

#Fig. 4. 1 IyK mot 4 SgP; SN som symboler og isolinjer for Maro_s ##Ikke relevant for siste gangs analyse

#sONV'er som skal plottes mot hverandre
s1 <- sONV_IyK
#s2 <- sONV_SgP

#Finner koordinatene for hj?rner i det 'plotteomr?det' med koordinater (x1, x2, y1, y2) som gir
#like enheter p? begge akser. Vi holder x1, x2 og y1 faste og bestemmer y2 slik at (x2-x1)/(y2-y1) i plt=c(x1,x2,y1,y2)
#blir lik max(s1)/max(s2)
#Det krever at y2 = [(x2-x1)*max(s2) + y1*max(s1)]/max(s1) 
#Setter vi x1, x2 og y1 konstante, f?r vi:
y2 <- (0.8*max(s2)+0.15*max(s1))/max(s1)
#y2

par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 4 SgP - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

polygon(c(0,0,0.8*max(s1),0.8*max(s1)),c(0,max(s2)/2,max(s2)/2,0),col="lightgray")
polygon(c(0,0,0.8*max(s1),0.8*max(s1)),c(0,max(s2)/2,max(s2)/2,0),col="lightgray",density=10) #legger egentlig et skr?tt raster opp?, og fjerner streken rundt

lines(c(0,max(s1)),c(max(s2)/2,max(s2)/2),lty=2,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0.025,max(s2)),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)),lty=2,col=8)

polygon(c(0,0,0.142,0.142),c(0,0.025,0.025,0),col="white")
polygon(c(0,0,0.142,0.142),c(0,0.025,0.025,0),col="white",density=10) #legger egentlig et skr?tt raster opp?, og fjerner streken rundt

lines(c(0,0.142),c(0,0),col=8)
lines(c(0,0.142),c(0.025,0.025),col=8)
lines(c(0,0),c(0,0.025),col=8)
lines(c(0.142,0.142),c(0,0.025),col=8)

points(0.01,0.012,pch=16,cex=0.5,col="black")
points(0.02,0.012,pch=16,cex=0.6,col="lightblue2")
points(0.03,0.012,pch=16,cex=0.75,col="darkblue")
points(0.04,0.012,pch=1,cex=0.75,col="red")
points(0.05,0.012,pch=16,cex=0.9,col="red")

text(0.06,0.012,"SN & Maro_s",adj=0,cex=0.8)
points(s1[SN==1],s2[SN==1],pch=16,cex=0.5,col=1)
points(s1[SN==2],s2[SN==2],pch=16,cex=0.6,col="lightblue2")
points(s1[SN==3],s2[SN==3],pch=16,cex=0.75,col="darkblue")
points(s1[SN==4],s2[SN==4],pch=1,cex=0.75,col="red")
points(s1[SN==5],s2[SN==5],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,4)],Maro_s,display="sites",levels=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
###

#Fig. 5. 1 IyK mot 4 SkP; SN som symboler og isolinjer for Arbla_a

#sONV'er som skal plottes mot hverandre
s1 <- sONV_IyK
s2 <- sONV_SkP

#Finner koordinatene for hj?rner i det 'plotteomr?det' med koordinater (x1, x2, y1, y2) som gir
#like enheter p? begge akser. Vi holder x1, x2 og y1 faste og bestemmer y2 slik at (x2-x1)/(y2-y1) i plt=c(x1,x2,y1,y2)
#blir lik max(s1)/max(s2)
#Det krever at y2 = [(x2-x1)*max(s2) + y1*max(s1)]/max(s1) 
#Setter vi x1, x2 og y1 konstante, f?r vi:
y2 <- (0.8*max(s2)+0.15*max(s1))/max(s1)
y2

par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 4 SkP - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

polygon(c(4*max(s1)/5,4*max(s1)/5,max(s1),max(s1)),c(0.025,max(s2),max(s2),0.025),col="lightgray")
polygon(c(4*max(s1)/5,4*max(s1)/5,max(s1),max(s1)),c(0.025,max(s2),max(s2),0.025),col="lightgray",density=10) #legger egentlig et skr?tt raster opp?, og fjerner streken rundt

polygon(c(max(s1)-0.142,max(s1)-0.142,4*max(s1)/5,4*max(s1)/5),c(0.025,max(s2)/2,max(s2)/2,0.025),col="lightgray")
polygon(c(max(s1)-0.142,max(s1)-0.142,4*max(s1)/5,4*max(s1)/5),c(0.025,max(s2)/2,max(s2)/2,0.025),col="lightgray",density=10) #legger egentlig et skr?tt raster opp?, og fjerner streken rundt

polygon(c(3*max(s1)/5,3*max(s1)/5,max(s1)-0.142,max(s1)-0.142),c(0,max(s2)/2,max(s2)/2,0),col="lightgray")
polygon(c(3*max(s1)/5,3*max(s1)/5,max(s1)-0.142,max(s1)-0.142),c(0,max(s2)/2,max(s2)/2,0),col="lightgray",density=10) #legger egentlig et skr?tt raster opp?, og fjerner streken rundt

lines(c(0,max(s1)),c(max(s2)/2,max(s2)/2),lty=2,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)

lines(c(max(s1)-0.142,max(s1)),c(0,0),col=8)
lines(c(max(s1)-0.142,max(s1)),c(0.025,0.025),col=8)
lines(c(max(s1)-0.142,max(s1)-0.142),c(0,0.025),col=8)
lines(c(max(s1),max(s1)),c(0,0.025),col=8)

points(max(s1)-0.142+0.01,0.012,pch=16,cex=0.5,col="black")
points(max(s1)-0.142+0.02,0.012,pch=16,cex=0.6,col="lightblue2")
points(max(s1)-0.142+0.03,0.012,pch=16,cex=0.75,col="darkblue")
points(max(s1)-0.142+0.04,0.012,pch=1,cex=0.75,col="red")
points(max(s1)-0.142+0.05,0.012,pch=16,cex=0.9,col="red")

text(max(s1)-0.142+0.06,0.012,"SN & Arbla_a",adj=0,cex=0.8)
points(s1[SN==1],s2[SN==1],pch=16,cex=0.5,col=1)
points(s1[SN==2],s2[SN==2],pch=16,cex=0.6,col="lightblue2")
points(s1[SN==3],s2[SN==3],pch=16,cex=0.75,col="darkblue")
points(s1[SN==4],s2[SN==4],pch=1,cex=0.75,col="red")
points(s1[SN==5],s2[SN==5],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,4)],Arbla_a,display="sites",levels=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
###

#Fig. 6. 1 IyK mot 6 OI; OI som symboler og isolinjer for Gab_a

#sONV'er som skal plottes mot hverandre
s1 <- sONV_IyK
s2 <- sONV_OI

#Finner koordinatene for hj?rner i det 'plotteomr?det' med koordinater (x1, x2, y1, y2) som gir
#like enheter p? begge akser. Vi holder x1, x2 og y1 faste og bestemmer y2 slik at (x2-x1)/(y2-y1) i plt=c(x1,x2,y1,y2)
#blir lik max(s1)/max(s2)
#Det krever at y2 = [(x2-x1)*max(s2) + y1*max(s1)]/max(s1) 
#Setter vi x1, x2 og y1 konstante, f?r vi:
y2 <- (0.8*max(s2)+0.15*max(s1))/max(s1)
y2


#Plotter OI-klasser p? figuren, og legger p? isolinjer for Gab_a
par(plt=c(0.15, 0.95, 0.15, y2))
plot(s1,s2,xlab="KLG 1 IYF - Coastal plains (KS457)",ylab="KLG 6 Gab_a - Coastal plains (KS457)",xlim=c(-0.01,max(s1)+0.01),ylim=c(-0.01,max(s2)+0.01),xaxp=c(0,0.4,5),yaxp=c(0,0.4,5),type="n") #Grunnlag for plotting av symboler etc.
lines(c(0,max(s1)),c(max(s2),max(s2)),lty=3,col=8)
lines(c(0,max(s1)),c(0,0),lty=3,col=8)
lines(c(0,0),c(0,max(s2)),lty=3,col=8)
lines(c(max(s1),max(s1)),c(0,max(s2)),lty=3,col=8)

lines(c(0,max(s1)),c(max(s2)/2,max(s2)/2),lty=2,col=8)
lines(c(0.2*max(s1),0.2*max(s1)),c(0,max(s2)-0.025),lty=2,col=8)
lines(c(0.4*max(s1),0.4*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.6*max(s1),0.6*max(s1)),c(0,max(s2)),lty=2,col=8)
lines(c(0.8*max(s1),0.8*max(s1)),c(0,max(s2)),lty=2,col=8)

lines(c(0,0.142),c(0,0),col=8)
lines(c(0,0.142),c(0.025,0.025),col=8)
lines(c(0.142,0.142),c(0,0.025),col=8)
lines(c(0,0),c(0,0.025),col=8)

points(0.01,0.012,pch=1,cex=0.6,col="lightblue3")
points(0.02,0.012,pch=16,cex=0.6,col="lightblue3")
points(0.03,0.012,pch=1,cex=0.75,col="lightblue4")
points(0.04,0.012,pch=16,cex=0.75,col="lightblue4")
points(0.05,0.012,pch=1,cex=0.9,col="red")
points(0.06,0.012,pch=16,cex=0.9,col="red")

text(0.07,0.012,"OI & Gab_a",adj=0,cex=0.8)
points(s1[OI==1],s2[OI==1],pch=1,cex=0.6,col="lightblue3")
points(s1[OI==2],s2[OI==2],pch=16,cex=0.6,col="lightblue3")
points(s1[OI==3],s2[OI==3],pch=1,cex=0.75,col="lightblue4")
points(s1[OI==4],s2[OI==4],pch=16,cex=0.75,col="lightblue4")
points(s1[OI==5],s2[OI==5],pch=1,cex=0.9,col="red")
points(s1[OI==6],s2[OI==6],pch=16,cex=0.9,col="red")

os <- ordisurf(snv[,c(1,6)],Gab_a,display="sites",levels=c(0.1,0.2,0.4,0.6,0.8), col="red",add=T) #inkluderer plotting p? eksisterende ordinasjonsdiagram
```


```{r}
library(goeveg)
library(vegan)

palette(c("#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", 
                    "#fdbf6f", "#ff7f00", "#cab2d6", "#6a3d9a", "#ffff99", "#b15928"))
#rgb( ramp(seq(0, 1, length = 5)), max = 255)

specresponse2 <- function(species, var, main, xlab, model = "auto", method = "env", axis = 1, points = FALSE, bw = FALSE) {
  
  if(!is.data.frame(species)) {
    
    if(missing(main)) {
      main <- deparse(substitute(species))
    }
    
    specnames <- deparse(substitute(species))
    species <- data.frame(species)
    ismat <- F
    
  } else {
    if(missing(main)) {
      main <- " "
    }
    specnames <- names(species)
    ismat <- T
  }
  
  species <- decostand(species, method = "pa")
  
  if(length(species) >= 1) {
    
    ls <- length(species)
    
    if(method == "env") {
      
      if(missing(xlab)) {
        xlab <- deparse(substitute(var))
      }
    } else if(method == "ord") {
      
      # Extract site scores from ordination
      var <- as.numeric(scores(var, display="sites", choices=axis))
      
      # X-axis labeling
      if(missing(xlab)) {
        xlab <- paste("Axis", axis, "sample scores")
      }
    } else {
      stop("Method unknown.")
    }
    
    plot(var, species[,1], main = main, type="n", frame = FALSE, 
         xlab = xlab, ylab="Probability of occurrence", ylim = c(0,1))
    
    for(i in 1:ls) {
      
      if(length(species[species[,i]>0,i]) <= 5) {
        # tried warning instead of print
        warning(paste("Only", length(species[species[,i]>0,i]), "occurrences of", names(species)[i], "."))
      }
      
      if(model == "unimodal") {
        
        specresponse <- suppressWarnings(glm(species[,i] ~ poly(var, 2),
                                             family="binomial"))
        
      } else if (model == "linear") {
        
        specresponse <- suppressWarnings(glm(species[,i] ~ var,
                                             family="binomial"))
        
      } else if (model == "bimodal") {
        
        specresponse <- suppressWarnings(glm(species[,i] ~ poly(var, 4),
                                             family="binomial"))
        
      }
      else if (model == "auto") {
        
        glm.1 <- suppressWarnings(glm(species[,i] ~ poly(var, 1), family="binomial"))
        glm.2 <- suppressWarnings(glm(species[,i] ~ poly(var, 2), family="binomial"))
        glm.3 <- suppressWarnings(glm(species[,i] ~ poly(var, 3), family="binomial"))
        glm.AIC <- c(extractAIC (glm.1)[2], extractAIC (glm.2)[2],
                     extractAIC (glm.3)[2])
        
        switch(c(1,2,3)[glm.AIC==min(glm.AIC)],
               {specresponse <- glm.1; deg<-1},
               {specresponse <- glm.2; deg<-2},
               {specresponse <- glm.3; deg<-3})
        
        #print(paste("GLM with", deg, "degrees fitted for", specnames[i]))
        
      } else if (model == "gam") {
        
        gam.list <- list()
        gam.AIC <- 0
        
        for(n in 1:4) {
          gam.list[[paste("gam", n, sep=".")]] <- mgcv::gam(species[,i] ~ s(var, k = n+2), family='binomial')
          gam.AIC[n] <- extractAIC(gam.list[[n]])[2]
        }
        
        switch(c(1,2,3,4)[gam.AIC==min(gam.AIC)],
               {specresponse <- gam.list[[1]]; deg<-3},
               {specresponse <- gam.list[[2]]; deg<-4},
               {specresponse <- gam.list[[3]]; deg<-5},
               {specresponse <- gam.list[[4]]; deg<-6})
        
        print(paste("GAM with", deg, "knots fitted for", specnames[i]))
        
      } else {
        stop("Model unknown.")
      }
      
      xneu <- seq(min(var), max(var), len = 101)
      preds <- predict(specresponse, newdata = data.frame(var = xneu),
                       type="response")
      
      if(bw == T) {
        if(points == TRUE) {
          if(i > 1) {
            species[,i][species[,i]==1] <- species[,i][species[,i]==1] - 0.02*(i-1)
            species[,i][species[,i]==0] <- species[,i][species[,i]==0] + 0.02*(i-1)
          }
          
          col <- col2rgb(i)
          points(var, species[,i], pch = i,
                 col = rgb( ramp(seq(0, 1, length = 5)), maxColorValue = 255))
        }
        
        lines(preds ~ xneu, lty=i)
      } else {
        
        if(points == TRUE) {
          if(i > 1) {
            species[,i][species[,i]==1] <- species[,i][species[,i]==1] - 0.02*(i-1)
            species[,i][species[,i]==0] <- species[,i][species[,i]==0] + 0.02*(i-1)
          }
          
          col <- col2rgb(i)
          points(var, species[,i],
                 col = rgb( ramp(seq(0, 1, length = 5)), maxColorValue = 255),
                 pch = 16)
        }
        
        lines(preds ~ xneu, lty = i, lwd= 2, col = i)
      }
    }
    
    if(ismat == T) {
      
      if(bw == T){
        if(points == TRUE) {
          legend(0.70, 0.75,legend= "names(species)", "b", lty=1:ls, lwd = 2, pch=1:ls,
                 bty = "n", cex = 0.85)
        } else {
          legend(0.70, 0.75, legend= "names(spees)", lty=1:ls, lwd = 2, pch=1:ls,
                 bty = "n", cex = 0.85)
        }
      } else {
        legend(0.70, 0.75, legend= c("Fishing-related buildings", "Built up area", "Town/city area", "Exposed coast", "Mire", "Technical heritage sites", "Steep coast", "Coniferous forest", "Impediment", "Open areas with heathland"), lty=1:ls, lwd = 2, col=1:ls, 
               bty = "n", cex = 0.85)
      }
    }
    
  }  else {
    stop("No species in matrix.")
  }
}
```


```{r}
par(mar=c(5.1, 4.1, 4.1, 12.1), xpd=TRUE)

specresponse2 (av[ ,c(2,3,31,44,32, 49,54,68,78,82)], main = " ", akse1, xlab = "GNMDS 1 (Coastal plains, outer- inner coast)")

#2,3,31,44,32, 49,54,68,78,82)



```

